---
title: "Facebook Birthdays"
output: html_document
---

# Your task is to investigate the distribution of your friends'
# birth months and days.

# Here some questions you could answer, and we hope you think of others.

# **********************************************************************

# How many people share your birthday? Do you know them?
# (Reserve time with them or save money to buy them a gift!)

# Which month contains the most number of birthdays?

# How many birthdays are in each month?

# Which day of the year has the most number of birthdays?

# Do you have at least 365 friends that have birthdays on everyday
# of the year?

# **********************************************************************

# You will need to do some data munging and additional research to
# complete this task. This task won't be easy, and you may encounter some
# unexpected challenges along the way. We hope you learn a lot from it though.

# You can expect to spend 30 min or more on this task depending if you
# use the provided data or obtain your personal data. We also encourage you
# to use the lubridate package for working with dates. Read over the documentation
# in RStudio and search for examples online if you need help.

# You'll need to export your Facebooks friends' birthdays to a csv file.
# You may need to create a calendar of your Facebook friends' birthdays
# in a program like Outlook or Gmail and then export the calendar as a
# csv file.

# Once you load the data into R Studio, you can use the strptime() function
# to extract the birth months and birth days. We recommend looking up the
# documentation for the function and finding examples online.

# We've included some links in the Instructor Notes to help get you started.

# Once you've completed your investigation, create a post in the discussions that includes:
#       1. any questions you answered, your observations, and summary statistics
#       2. snippets of code that created the plots
#       3. links to the images of your plots

# You can save images by using the ggsave() command.
# ggsave() will save the last plot created.
# For example...
#                  qplot(x = price, data = diamonds)
#                  ggsave('priceHistogram.png')

# ggsave currently recognises the extensions eps/ps, tex (pictex),
# pdf, jpeg, tiff, png, bmp, svg and wmf (windows only).

# Copy and paste all of the code that you used for
# your investigation below the line. Submit it when you are ready.
# ===============================================================================

```{r}
setwd("~/Udacity/dataAnalysisR/EDA_R_Udacity/lesson3")
library(rmarkdown)
library(ggplot2)
library(lubridate)
```

#Investigating the Distribution of Friend's Birth Months and Days#

 Which month contains the most number of birthdays?

-How many birthdays are in each month?
-Which day of the year has the most number of birthdays?
-Do you have at least 365 friends that have birthdays on everyday of the year?
-How many people share my birthday?

##Read in the data##

This first part is from the following example(http://rpubs.com/Daria/52453).  I read about lubridate, I read about handling dates in R.  In the end I was more comfortable taking an approach like I would in PANDAS.

```{r}
pf <- read.csv('birthdaysExample.csv')
names(pf)
```
```{r}
#convert vector into datetime format
bdates <- strptime(pf$dates, "%m/%d/%y")
names(unclass(bdates))
head(bdates)
```

```{r}
#extract days and months
days <- bdates$mday
mon <- as.numeric(bdates$mon) + 1 
yday <- as.numeric(bdates$yday) + 1

#create dataframes
df <- data.frame(days, mon, yday, count = 1)
head(df)
```

