```{r}
getwd()
list.files()
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
names(pf)
```


```{r}
library('ggplot2')
qplot(x = dob_day, data = pf) +
  scale_x_discrete(breaks = 1:31)


```
```{r}
#install.packages('ggthemes', dependencies = TRUE) 
library(ggthemes) 
```

```{r}
ggplot(aes(x = dob_day), data = pf) + 
  geom_histogram() + 
  scale_x_discrete(breaks = 1:31)

```
```{r}
ggplot(aes(x = dob_day), data = pf) + 
  geom_histogram() + 
  scale_x_discrete(breaks = 1:31) +
  facet_wrap(~dob_month,ncol = 3)
```
```{r}
summary(pf)
names(pf)
```

```{r}
ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram()  
  #+ scale_x_discrete(breaks = 1:31) +
  #facet_wrap(~dob_month,ncol = 3)
```

```{r}
ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram() + 
  scale_x_continuous(limits = c(0, 1000))
```
```{r}
ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram(binwidth = 25) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50))
```

```{r}
qplot(x = friend_count, data = pf, binwidth = 25) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```
```{r}
ggplot(aes(x = friend_count), data = subset(pf, !is.na(gender))) + 
  geom_histogram() + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) + 
  facet_wrap(~gender)
```

```{r}
table(pf$gender)
```
```{r}
by(pf$friend_count, pf$gender, summary)
```
```{r}
ggplot(aes(x = tenure), data = pf) + 
   geom_histogram(binwidth = 30, color = 'black', fill = '#099DD9')
```

```{r}
ggplot(aes(x = tenure/365), data = pf) + 
   geom_histogram(binwidth = 1, color = 'black', fill = '#099DD9')
```
```{r}
#instructor ggplot solution see notes for qplot solution
ggplot(aes(x = tenure/365), data = pf) + 
   geom_histogram(binwidth = .25, color = 'black', fill = '#F79420')
```


```{r}
#I set a new scale to zoom in on the data
ggplot(aes(x = tenure/365), data = pf) + 
   geom_histogram(binwidth = .25, color = 'black', fill = '#F79420') + 
   scale_x_continuous(limits = c(0, 7), breaks = seq(1,7,1))
```
```{r}
#add labels
ggplot(aes(x = tenure / 365), data = pf) + 
  geom_histogram(binwidth = .25,color = 'black', fill = '#F79420') + 
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7)) + 
  xlab('Number of years using Facebook') + 
  ylab('Number of users in sample')
```
```{r}
#quiz to create a histogram of ages and play with data
#what names do we have to work with?
names(pf)
summary(pf$age)
#now to create a histogram of ages
ggplot(aes(x = age), data = pf) + 
  geom_histogram(binwidth = 1,color = 'black', fill = '#F79420') + 
  xlab('Age') + 
  ylab('Frequency')
#hmm, the upper extremes strain credulity
  
```

```{r}
#let's split this up a bit
ggplot(aes(x = age), data = pf) + 
  geom_histogram(binwidth = 5,color = 'black', fill = '#F79420') + 
  scale_x_continuous(breaks = seq(5, 110, 5), limits = c(10, 110)) +
  xlab('Age') + 
  ylab('Frequency')
```

```{r}
#let's look at that upper range
ggplot(aes(x = age), data = pf) + 
  geom_histogram(binwidth = 10,color = 'black', fill = '#F79420') + 
  scale_x_continuous(breaks = seq(10, 110, 10), limits = c(90, 110)) +
  xlab('Age') + 
  ylab('Frequency')
```
```{r}
#install.packages('gridExtra') 
library(gridExtra) 
```

```{r}
#create 1 column with 3 histograms: friend count, log10 friend count, sqrt friend #count

p1 = ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram(binwidth = 25, fill = '#F79420') + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50))

p2 = ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram(fill = '#F79420') + 
  scale_x_log10() +
  xlab('log10(friend_count)')

p3 = ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram(fill = '#F79420') + 
  scale_x_sqrt() +
  xlab('sqrt(friend_count)')


# arrange plots in grid
grid.arrange(p1, p2, p3, ncol=1)
```
```{r}
#Here is a better way of doing it
p1 = ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram(fill = '#F79420')

p2 = p1 + scale_x_log10() +
  xlab('log10(friend_count)')
  
p3 = p1 +
  scale_x_sqrt() +
  xlab('sqrt(friend_count)')

grid.arrange(p1,p2,p3,ncol=1)

```

```{r}
#Use a frequency polygon to determine which gender makes more likes on the www (www_likes)

#we need to subset the data for not gender = na and add y ..etc to get proportion. The latter is because there are many more males than females in the data set.  We know from previous analysis that females tend to have more likes, where does the transitions occur?

ggplot(aes(x = friend_count, y = ..count../sum(..count..)), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender)) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) + 
  xlab('Friend Count') + 
  ylab('Percentage of users with that friend count')
```
```{r}
#changing plot range

ggplot(aes(x = friend_count, y = ..count../sum(..count..)), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender)) + 
  scale_x_continuous(limits = c(530, 600), breaks = seq(530, 600, 5)) + 
  xlab('Friend Count') + 
  ylab('Percentage of users with that friend count')

```
```{r}
#I need to do some more work:  - www_likes is a count variable and should be plotted on a discrete scale  - www_like is long tail data which may benefit from a log tranformation

ggplot(aes(x = www_likes), data = subset(pf, !is.na(gender))) + 
  geom_freqpoly(aes(color = gender)) + 
  scale_x_log10()
```

**test markdown outside r block:**
```{r}
#let's look at this numerically

sum(subset(pf$www_likes, pf$gender == 'male'))

```
**compare means**
```{r}
by(pf$www_likes, pf$gender, summary)

```
**compare sum**
```{r}
by(pf$www_likes, pf$gender, sum)
```


**look at the means, females have more likes**

##Back to Friend Counts##

**Adjust code to:**
- focus on users in population sample who have friend counts between 0 and 1000.
- going to try this with qplot and ggplot

###basic qplot###

```{r}
qplot(x = gender, y = friend_count,data = subset(pf,!is.na(gender)),geom = 'boxplot')
```
###wow that is a lot of outliers (the dots)###

**let's try to zoom in closer to the IQR**

```{r}
qplot(x = gender, y = friend_count,data = subset(pf,!is.na(gender)),geom = 'boxplot',ylim = c(0,1000))
```
**so it turns out that ylim removes data points, to get around this I will add a scale_y_continuous layer**

```{r}
qplot(x = gender, y = friend_count,data = subset(pf,!is.na(gender)),geom = 'boxplot') +
  scale_y_continuous(limits = c(0,1000))
```
**ZOMG! this removes points too, one more time with coord_cartesian()**

```{r}
qplot(x = gender, y = friend_count,data = subset(pf,!is.na(gender)),geom = 'boxplot') +
  coord_cartesian(ylim = c(0,1000))
  
```
**ok good, that did not remove data, now to replicate in ggplot**
**update decided to move on to other things and circle back**

**median (black line) slightly higher for females, let's look more closely**

```{r}
qplot(x = gender, y = friend_count,data = subset(pf,!is.na(gender)),geom = 'boxplot') +
  coord_cartesian(ylim = c(0,250))
```
**yeah, slightly higher**

###Does this match our tables?###

```{r}
by(pf$friend_count,pf$gender,summary)
```
**yes it does!**
####What have we observed?####
- 25% of femaile users have > 244 friends vs. males where 25% have > 182
- Median higer for females than males

**So we know this now, but what we don't know is who actually initiated more friendships?  Who is more active in trying to obtain friends**

**What do we have to choose from?**

```{r}
names(pf)
```
**Looks like friendships initiated is the way to go**

```{r}
qplot(x = gender, y = friendships_initiated,data = subset(pf,!is.na(gender)),geom = 'boxplot') +
  coord_cartesian(ylim = c(0,150))
```
```{r}
by(pf$friendships_initiated,pf$gender,summary)
```
**And the females have it ladies and gentlemen!**

