---
title: "problemSet3"
output: html_document
---

#GGPlot2 Diamonds Data Set##

We need to perform some analysis on the ggplot2 diamonds data set to answer some questions.

**Questions:**
- How many observations are in the data set?
- How many variables are in the data set?
- How many ordered factors are in the data set?
- What letter represents the best diamond color?

##Load and Examine the data##

```{r}
library(ggplot2)
data(diamonds)
summary(diamonds)
```
There are 53940 observations and 10 variables.
Ordered factors inculude: cut, color, and clarity

color ranks include D, E, F, G, H, I, J

Which is best, let's look at the included documentation.

```{r}
?diamonds
```
>colour. diamond colour, from J (worst) to D (best).  So, D is the best.

##Create a Histogram of Diamond Prices##

```{r}
library(ggthemes)
```
```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = 100,  color = 'black',fill = '#F79420')
```
Data is long-tailed, transforming onto log scale

```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = .05,  color = 'black',fill = '#F79420') + 
  scale_x_log10()+
  xlab('Log10(price)')
  
```
##Descriptive Statistics of Price##
```{r}
summary(diamonds$price)
```
##Diamond Counts#

How many diamonds cost less than 500 USD?

```{r}
nrow(subset(diamonds, diamonds$price < 500))
```



How many diamonds cost less than 250 USD?
```{r}
nrow(subset(diamonds, diamonds$price < 250))
```

How many diamonds at least 15,000 USD?

```{r}
nrow(subset(diamonds, diamonds$price >= 15000))
```





