---
title: "problemSet3"
output: html_document
---

#GGPlot2 Diamonds Data Set##

We need to perform some analysis on the ggplot2 diamonds data set to answer some questions.

**Questions:**
- How many observations are in the data set?
- How many variables are in the data set?
- How many ordered factors are in the data set?
- What letter represents the best diamond color?

##Load and Examine the data##

```{r}
library(ggplot2)
data(diamonds)
summary(diamonds)
```
There are 53940 observations and 10 variables.
Ordered factors inculude: cut, color, and clarity

color ranks include D, E, F, G, H, I, J

Which is best, let's look at the included documentation.

```{r}
?diamonds
```
>colour. diamond colour, from J (worst) to D (best).  So, D is the best.

##Create a Histogram of Diamond Prices##

```{r}
library(ggthemes)
```
```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = 100,  color = 'black',fill = '#F79420')
```
Data is long-tailed, transforming onto log scale

```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = .05,  color = 'black',fill = '#F79420') + 
  scale_x_log10()+
  xlab('Log10(price)')
  
```
The data appears to be bimodal.

##Descriptive Statistics of Price##
```{r}
summary(diamonds$price)
```
##Diamond Counts#

How many diamonds cost less than 500 USD?

```{r}
nrow(subset(diamonds, diamonds$price < 500))
```



How many diamonds cost less than 250 USD?
```{r}
nrow(subset(diamonds, diamonds$price < 250))
```

How many diamonds at least 15,000 USD?

```{r}
nrow(subset(diamonds, diamonds$price >= 15000))
```

##Eplore the largest Price Bin##

```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = 200,  color = 'black',fill = '#F79420') +
  scale_x_continuous(limits = c(250, 1000), breaks = seq(250, 1000, 200))

```

```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = 100,  color = 'black',fill = '#F79420') +
  scale_x_continuous(limits = c(250, 1000), breaks = seq(250, 1000, 50))
```

```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = 25,  color = 'black',fill = '#F79420') +
  scale_x_continuous(limits = c(500, 1000), breaks = seq(500, 1000, 25))
```

The last one is too granular.  600 to 800 USD seems to be the sweet spot for low end diamond prices.  

**Insight provide from the class:**

>There are no diamonds that cost $1500.
>For diamonds that cost less than $2,000, the most common price of a diamond is >around $700 with the mode being $605 (binwidth = 1). 

I would like to explore this.

```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = 1,  color = 'black',fill = '#F79420') +
  scale_x_continuous(limits = c(600, 850), breaks = seq(600, 850, 10))
```
```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = 1,  color = 'black',fill = '#F79420') +
  scale_x_continuous(limits = c(600, 610), breaks = seq(600, 610, 1))
```
yup, 605.

```
```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = 100,  color = 'black',fill = '#F79420') +
  scale_x_continuous(limits = c(1000, 2000), breaks = seq(1000, 2000, 100))
```
There has to be a better way to find disconinuities, but I'll keep zooming for now.

```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = 10,  color = 'black',fill = '#F79420') +
  scale_x_continuous(limits = c(1400, 1600), breaks = seq(1400, 1600, 10))
```

Yeah, there is a large discontinuity.


##Break out the Histogram Diamond Prices by Cut##

I could use a grid to do this, but I'm going to start by trying a facet wrap.

```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = 100,  color = 'black',fill = '#F79420') +
  facet_wrap(~cut,ncol = 1)
```
Data is long-tailed, transforming onto log scale

```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = .05,  color = 'black',fill = '#F79420') + 
  scale_x_log10()+
  xlab('Log10(price)') +
  facet_wrap(~cut,ncol = 1)
```

The distributions have similiar shapes, the most different is the fair cut.

##Basic Price Information by Cut##
```{r}
by(diamonds$price,diamonds$cut,max)
by(diamonds$price,diamonds$cut,min)
by(diamonds$price,diamonds$cut,median)
```
##Free Scales##

```{r}
# In the two last exercises, we looked at
# the distribution for diamonds by cut.

# Run the code below in R Studio to generate
# the histogram as a reminder.

# ===============================================================

qplot(x = price, data = diamonds) + facet_wrap(~cut)



                                                                               
# ===============================================================

# In the last exercise, we looked at the summary statistics
# for diamond price by cut. If we look at the output table, the
# the median and quartiles are reasonably close to each other.

# diamonds$cut: Fair
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#     337    2050    3282    4359    5206   18570 
# ------------------------------------------------------------------------ 
# diamonds$cut: Good
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#     327    1145    3050    3929    5028   18790 
# ------------------------------------------------------------------------ 
# diamonds$cut: Very Good
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#     336     912    2648    3982    5373   18820 
# ------------------------------------------------------------------------ 
# diamonds$cut: Premium
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#     326    1046    3185    4584    6296   18820 
# ------------------------------------------------------------------------ 
# diamonds$cut: Ideal
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#     326     878    1810    3458    4678   18810 

# This means the distributions should be somewhat similar,
# but the histograms we created don't show that.

# The 'Fair' and 'Good' diamonds appear to have 
# different distributions compared to the better
# cut diamonds. They seem somewhat uniform
# on the left with long tails on the right.

# Let's look in to this more.

# Look up the documentation for facet_wrap in R Studio.
# Then, scroll back up and add a parameter to facet_wrap so that
# the y-axis in the histograms is not fixed. You want the y-axis to
# be different for each histogram.

# If you want a hint, check out the Instructor Notes.
```

```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = 100,  color = 'black',fill = '#F79420') +
  facet_wrap(~cut,ncol = 1, scales = 'free')
```


```{r}
ggplot(aes(x = price), data = diamonds) + 
  geom_histogram(binwidth = .05,  color = 'black',fill = '#F79420') + 
  scale_x_log10()+
  xlab('Log10(price)') +
  facet_wrap(~cut,ncol = 1, scales = 'free')
```

Now what we visualize is mathing up with what we know about the data (similiar distributions)

##Histogram of Price per Carat Faceted by Cut##

```{r}
ggplot(aes(x = price/carat), data = diamonds) + 
  geom_histogram(binwidth = 100,  color = 'black',fill = '#F79420') +
  facet_wrap(~cut,ncol = 1, scales = 'free')
```


```{r}
ggplot(aes(x = price/carat), data = diamonds) + 
  geom_histogram(binwidth = .05,  color = 'black',fill = '#F79420') + 
  scale_x_log10()+
  xlab('Log10(price/carat)') +
  facet_wrap(~cut,ncol = 1, scales = 'free')
```

##Investigate the price of diamonds using box plots,numerical summaries, and one of the following categorical variables: cut, clarity, or color.##

```{r}
library(gridExtra) 
```
```{r}
clarity = qplot(x = clarity, y = price,data = diamonds,geom = 'boxplot') +
  coord_cartesian()
  

color = qplot(x = color, y = price,data = diamonds,geom = 'boxplot') +
  coord_cartesian()

cut = qplot(x = cut, y = price,data = diamonds,geom = 'boxplot') +
  coord_cartesian()

grid.arrange(clarity, color, cut, ncol = 1)
```

```{r}
clarity = qplot(x = clarity, y = price,data = diamonds,geom = 'boxplot') +
  coord_cartesian(ylim = c(250,7000))
  

color = qplot(x = color, y = price,data = diamonds,geom = 'boxplot') +
  coord_cartesian(ylim = c(250,7000))

cut = qplot(x = cut, y = price,data = diamonds,geom = 'boxplot') +
  coord_cartesian(ylim = c(250,7000))

grid.arrange(clarity, color, cut, ncol = 1)
```
```{r}
by(diamonds$price,diamonds$cut, median)
by(diamonds$price,diamonds$cut, max)
by(diamonds$price,diamonds$cut, min)
by(diamonds$price,diamonds$cut, mean)
by(diamonds$price,diamonds$cut, median) - by(diamonds$price,diamonds$cut, mean)
```


I just find the trends in the medians interesting.  For cut, the medians are similiar except for ideal.  For color they tend to go up in a linear fashion to J (worst color rating), and for clarity they tend to go down.  What I see is that there are a variety of factors tha affect price. I think to really understand this we have to plot and view the data in other ways.


##Numerical Analysis##

-What is the price range for the middle 50% of diamonds with color D?

```{r}
by(diamonds$price,(diamonds$color == 'D'), summary)
```



-What is the price range for the middle 50% of diamonds with color J?

```{r}
by(diamonds$price,(diamonds$color == 'J'), summary)
```


-What is the IQR for diamonds wit the best color?

```{r}
by(diamonds$price,(diamonds$color == 'D'), IQR)
```


-What is the IQR for diamonds with the worst color?

```{r}
by(diamonds$price,(diamonds$color == 'J'), IQR)
```

Instructor Notes

You can use the function IQR() to find the interquartile range. Pass it a subset of the diamonds data frame. 

For example... 
IQR(subset(diamonds, price <1000)$price) 
Remember subset returns a data frame so we need to use $price on the end to access that variable.


##Investigate the price per carat of diamonds across the different colors of diamonds using boxplots.##

```{r}
qplot(x = color, y = price/carat,data = diamonds,geom = 'boxplot') +
  coord_cartesian()

qplot(x = color, y = price/carat,data = diamonds,geom = 'boxplot') +
  coord_cartesian(ylim = c(250,7000))
```

##Investigate weight of diamonds (carat) with a frequency polygon##

**how does binwidth affect the polygon?**

```{r}
ggplot(aes(x = carat), data = diamonds) + 
  geom_freqpoly(binwidth = .5,color = 'red')+
  scale_x_log10()

ggplot(aes(x = carat), data = diamonds) + 
  geom_freqpoly(binwidth = .05,color = 'red')+
  scale_x_log10()

ggplot(aes(x = carat), data = diamonds) + 
  geom_freqpoly(binwidth = .01,color = 'red')+
  scale_x_log10()
```
**What carat size(s) has a count > 2000?**

```{r}
#I have been using subset to do this, but instead used brackets as a select
#statement here, much like I do in PANDAS.
table(diamonds$carat)[table(diamonds$carat)>2000]
```


