Lesson 4
========================================================
```{r 'setwd'}
setwd("~/Udacity/dataAnalysisR/EDA_R_Udacity/lesson4")
```
###Load Library and Data Sets###

```{r 'load libs,data'}
library(ggplot2)
library(gridextra)
getwd()
list.files()
pf <- read.csv("~/Udacity/dataAnalysisR/EDA_R_Udacity/lesson3/pseudo_facebook.tsv", sep = '\t')
names(pf)
#alternative way to read file read.delim('pseudo_facebook.tsv')
```

***

### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:

```{r Scatterplots}
#qplot syntax
qplot(age,friend_count,data =pf)

#ggplot syntax
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point()
```

***

#### What are some things that you notice right away?
Response:  Younger people seem to have more friends.  There are some people of advanced age with a lot of friends.  I'm guessing these are not their actual biological age.  

***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
#qplot syntax
qplot(age,friend_count,data =pf)

#ggplot syntax
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point() +
  xlim(13,90) #min facebook age to reasonable max age

summary(pf$age)
```

***

### Overplotting
Notes:  We have so many points that it is difficult to see how many points are plotted, especially in the under 30 region.  We can use a transparency layer (alpha) to help with this.

```{r Overplotting}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20) + #alpha layer takes 20 points to make a dot completely black
  xlim(13,90) + #min facebook age to reasonable max age
  
#age is a continuous function but we have stacked columns (discrete), add jitter noise (dispersion) help us getter a better feel for the relationship between age and friend count.
  
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_jitter(alpha = 1/20) + #alpha layer takes 20 points to make one black dot
  xlim(13,90)  #min facebook age to reasonable max age
  
```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:

```{r Coord_trans()}

```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}

```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}

```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

```{r Conditional Means}

```

Create your plot!

```{r Conditional Means Plot}

```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}

```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}

```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(                 , cor.test(age, friend_count))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}

```

***

### Strong Correlations
Notes:

```{r Strong Correlations}

```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}

```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
install.packages('alr3')
library(alr3)
```

Create your plot!

```{r Temp vs Month}

```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}

```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}

```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}

```

***

### Age with Months Means

```{r Age with Months Means}

```

Programming Assignment
```{r Programming Assignment}

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}

```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}

```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

