Lesson 4
========================================================
```{r 'setwd'}
setwd("~/Udacity/dataAnalysisR/EDA_R_Udacity/lesson4")
```
###Load Library and Data Sets###

```{r 'load libs,data'}
library(ggplot2)
library(gridExtra)
library(dplyr)
library('alr3')
getwd()
list.files()
pf <- read.csv("~/Udacity/dataAnalysisR/EDA_R_Udacity/lesson3/pseudo_facebook.tsv", sep = '\t')
names(pf)
#alternative way to read file read.delim('pseudo_facebook.tsv')
```

***

### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:

```{r Scatterplots}
#qplot syntax
qplot(age,friend_count,data =pf)

#ggplot syntax
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point()
```

***

#### What are some things that you notice right away?
Response:  Younger people seem to have more friends.  There are some people of advanced age with a lot of friends.  I'm guessing these are not their actual biological age.  

***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
#qplot syntax
qplot(age,friend_count,data =pf)

#ggplot syntax
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point() +
  xlim(13,90) #min facebook age to reasonable max age

summary(pf$age)
```

***

### Overplotting
Notes:  We have so many points that it is difficult to see how many points are plotted, especially in the under 30 region.  We can use a transparency layer (alpha) to help with this.

```{r Overplotting}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20) + #alpha layer takes 20 points to make a dot completely black
  xlim(13,90) #min facebook age to reasonable max age
  
#age is a continuous function but we have stacked columns (discrete), add jitter noise (dispersion) help us getter a better feel for the relationship between age and friend count.
  
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_jitter(alpha = 1/20) + #alpha layer takes 20 points to make one black dot
  xlim(13,90)  #min facebook age to reasonable max age
  
```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:

```{r Coord_trans()}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point() +
  xlim(13,90) +
  coord_trans(y = 'sqrt')
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20) +
  xlim(13,90) +
  coord_trans(y = 'sqrt')
```

#### What do you notice?
I noticed that this is not a helpful transformation.
***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
ggplot(aes(x = age, y = friendships_initiated), data = pf) + 
  geom_point()

ggplot(aes(x = age, y = friendships_initiated), data = pf) + 
  geom_jitter(alpha = 1/20) + #jitter for dispersion appropropriate to continuous var of age
  xlim(13,90)

ggplot(aes(x = age, y = friendships_initiated), data = pf) + 
  geom_point(alpha = 1/20) + #switch to geom point to avoid -/i numbers
  xlim(13,90) +
  coord_trans(y = 'sqrt')

ggplot(aes(x = age, y = friendships_initiated), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h = 0)) + #use position and limit to get jitter dispersion without -/i #s
  xlim(13,90) +
  coord_trans(y = 'sqrt')




```

***

### Overplotting and Domain Knowledge
Notes:  Normalizing audience size to % simplified the data.

***

### Conditional Means
Notes:  There are a lot of great links to dplyr infor as well as alternative functions to dplyr, such as split and lapply.

```{r Conditional Means}
#group_by and some other agg functions using dplyr
#buld the dataframe with a dplyr agg function
age_groups <- group_by(pf,age)
#build the table
pf.fc_by_age <- summarise(age_groups, 
                          friend_count_mean = mean(friend_count),
                          friend_count_median = median(friend_count),
                          n = n()) #count function

pf.fc_by_age = arrange(pf.fc_by_age, age) #sort by age

head(pf.fc_by_age)

#an alternative way of creatig the same table by chaining functions onto dataset
pf.fc_by_age <- pf %.%
  group_by(age) %.%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %.%
  arrange(age)

head(pf.fc_by_age,20)
  
  
  
```

Create your plot!

```{r Conditional Means Plot}
ggplot(aes(x = age, y = friend_count_mean), data = pf.fc_by_age) + 
  geom_line() +
  xlim(13,90) 

```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x = age, y = friend_count), data = pf) +
  coord_cartesian(xlim = c(13, 90), ylim = c(0,1000)) +
  geom_point(alpha = 1/20, position = position_jitter(h = 0),
             color = 'orange') + #use position and limit to get jitter dispersion without -/i #s
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, probs = .1,
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = .5,
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = .9,
            linetype = 2, color = 'blue') 

ggplot(aes(x= age, y= friend_count), data = pf) +
  coord_cartesian(xlim = c(13, 31)) +
  geom_point(alpha = 1/20, position = position_jitter(h = 0),
             color = 'orange') + 
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, probs = .1,
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = .5,
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = .9,
            linetype = 2, color = 'blue')
  
```

#### What are some of your observations of the plot?
Response: the 90th percentile is below 800 friends.

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
cor.test(pf$age, pf$friend_count, method = c('pearson'))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response: -.027

***

### Correlation on Subsets
Notes:  We should eliminate the higher ages, since we reasonably believe them to be erroneous.

```{r Correlation on Subsets}
with(subset(pf, pf$age <= 70), cor.test(age, friend_count))
```
This produced a weaker negative correlation (as age decreases, number of friends decreases.  This does not demonstrate causality, for that we would need to use inferential vs. descriptive statistics.)
***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
# Create a scatterplot of likes_received (y)
# vs. www_likes_received (x). Use any of the
# techniques that you've learned so far to
# modify the plot.
ggplot(aes(x= www_likes_received, y= likes_received), data = pf) +
  geom_point()

#lets zoom in on the most dense area
ggplot(aes(x= www_likes_received, y= likes_received), data = pf) +
  coord_cartesian(xlim = c(0,5000), ylim = c(-100, 5000)) +
  geom_point()

#hard to see the very dense region
ggplot(aes(x= likes_received, y= www_likes_received), data = pf) +
  coord_cartesian(xlim = c(0,2000), ylim = c(-300, 2000)) +
  geom_point(alpha = .05, position = position_jitter(h = 0),
             color = 'orange')


```

***

### Strong Correlations
Notes:

```{r Strong Correlations}
ggplot(aes(x= www_likes_received, y= likes_received), data = pf) +
  geom_point() +
  xlim(0, quantile(pf$www_likes_received,0.95)) +
  ylim(0,quantile(pf$likes_received,0.95)) +
  geom_smooth(method = 'lm',color = 'red')
```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
with(pf, cor.test(www_likes_received, likes_received))
```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
#install.packages('alr3')
library(alr3)
data(Mitchell)
?Mitchell
head(Mitchell)
```

Create your plot!

```{r Temp vs Month}
ggplot(aes(x= Month, y= Temp), data = Mitchell) +
  geom_point()
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

.1

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}
with(Mitchell,cor.test(Month, Temp))
```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}
ggplot(aes(x= Month, y= Temp), data = Mitchell) +
  geom_point() +
  scale_x_discrete(breaks = seq(0,203,12)) #found months by typing range(Mitchell$Month) directly into console

#Let's stretch out the horizontal axis
ggplot(aes(x= Month, y= Temp), data = Mitchell) +
  geom_point() +
  scale_x_discrete(breaks = seq(0,203,12))

ggplot(aes(x=(Month%%12),y=Temp),data=Mitchell)+ 
  geom_point() 
```
Instructor Notes

You could also get perspective on this data by overlaying each year's data on top of each other, giving a clear, generally sinusoidal graph. You can do this by using the R's modulus operator %% in your code. Try running the code below!

ggplot(aes(x=(Month%%12),y=Temp),data=Mitchell)+ 
  geom_point() 

Data Visualization Pioneers 
John Tukey 
William Playfair 
William Playfair and the Psychology of Graphs 

There are other measures of associations that can detect this. The dcor.ttest() function in the energy package implements a non-parametric test of the independence of two variables. The test correctly rejects the independence.
***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
pf$age_with_months <- pf$age + (1-pf$dob_month/12)
```

***

### Age with Months Means

```{r Age with Months Means}

```

Programming Assignment
```{r Programming Assignment}
# Create a new data frame called
# pf.fc_by_age_months that contains
# the mean friend count, the median friend
# count, and the number of users in each
# group of age_with_months. The rows of the
# data framed should be arranged in increasing
# order by the age_with_months variable.

# For example, the first two rows of the resulting
# data frame would look something like...

# age_with_months  friend_count_mean  friend_count_median	n
#              13            275.0000                   275 2
#        13.25000            133.2000                   101 11

pf.fc_by_age_months <- pf %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age_with_months)

diamondsByClarity
```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}
# Create a new scatterplot showing friend_count_mean
# versus the new variable, age_with_months. Be sure to use
# the correct data frame (the one you create in the last
# exercise) AND subset the data to investigate
# users with ages less than 71.

ggplot(aes(x= age_with_months, y= friend_count_mean), data = subset(pf.fc_by_age_months,pf.fc_by_age_months$age_with_months < 71)) +
  geom_point() +
  geom_smooth(method = 'lm',color = 'red')

p1 = ggplot(aes(x= age, y= friend_count_mean), data = subset(pf.fc_by_age,pf.fc_by_age$age < 71)) +
  geom_line() +
  geom_smooth(method = 'lm',color = 'red')


p2 = ggplot(aes(x= age_with_months, y= friend_count_mean), data = subset(pf.fc_by_age_months,pf.fc_by_age_months$age_with_months < 71)) +
  geom_line() +
  geom_smooth(method = 'lm',color = 'red')




p3 = ggplot(aes(x= round(age/5)*5, y= friend_count), data = subset(pf,pf$age < 71)) +
  geom_line(stat = 'summary', fun.y = mean) + #plotting mean of friend count rather than friend count
  geom_smooth(method = 'lm',color = 'red')


grid.arrange(p2,p1,p3,ncol=1)

p4 = ggplot(aes(x= age, y= friend_count_mean), data = subset(pf.fc_by_age,pf.fc_by_age$age < 71)) +
  geom_line() +
  geom_smooth()


p5 = ggplot(aes(x= age_with_months, y= friend_count_mean), data = subset(pf.fc_by_age_months,pf.fc_by_age_months$age_with_months < 71)) +
  geom_line() +
  geom_smooth()


p6 = ggplot(aes(x= round(age/5)*5, y= friend_count), data = subset(pf,pf$age < 71)) +
  geom_line(stat = 'summary', fun.y = mean) + #plotting mean of friend count rather than friend count
  geom_smooth()

grid.arrange(p5,p4,p6,ncol=1)
```
Looking at the 3 plots the top one (p2) has more bins (age in months), the middle plot (p1, age in years) has less bins, it is smoother.  For plot 3 we lump everyone together whose age falls on a multiple of five.  For example, users who are within 2.5 years of 40, 50, and 60 will be lumped together, respectively.
Look at Local Regression (LOESS) explained visually on the Simply Statistics blog.  Now we have less data points and wider bin widths.  We may now be able to estimate more accurately but have lost some of the interesting features of the age and friend_count relationship.  This is bias variance tradeoff.  A better way to handle the noise is to use a smooth function.  Using these functions, splines etc. makes the assumption that the true data is smooth.  In this case it isn't, so while it does smooth out the noise, we miss the nonmontonic features of the low age ranges and the spike at age 69.

The Details of Loess and Lowess
***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}

```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

