---
title: "Lesson4_problemset"
output: html_document
---

```{r 'setwd'}
setwd("~/Udacity/dataAnalysisR/EDA_R_Udacity/lesson4")
```
###Load Library and Data Sets###

```{r 'load libs,data'}
library(ggplot2)
library(gridExtra)
library(dplyr)
library('alr3')
library(dplyr)
getwd()
list.files()
data(diamonds)
summary(diamonds)
names(diamonds)
```

```{r scatterplot price vs. x}
ggplot(aes(x = x , y = price), data = diamonds ) +
  geom_point(color = 'orange')

```
```{r correlation of price}
cor.test(diamonds$x, diamonds$price)
cor.test(diamonds$y,diamonds$price)
cor.test(diamonds$z, diamonds$price)

```
```{r price vs. depth}
ggplot(aes(x = depth, y = price), data = diamonds) +
  geom_point(color = 'red')
```

```{r adjustments price vs. depth}
ggplot(aes(x = depth, y = price), data = diamonds) +
  geom_point(alpha = .01, color = 'red') +
  scale_x_continuous(breaks = seq(40,80,2))
```

```{r correlation depth vs. price}
cor.test(diamonds$depth, diamonds$price)
```

```{r price vs. carat}
ggplot(aes(x = carat, y = price,), data = diamonds)+
  geom_point(color = 'blue') +
  xlim(0, quantile(diamonds$carat,0.99)) +
  ylim(0, quantile(diamonds$price,0.99))
```
```{r price vs. volume}
volume = diamonds$x * diamonds$y * diamonds$z
ggplot(aes(x = volume, y = price), data = diamonds) +
  geom_point(color = 'green')
```
```{r correlations on subsets}
diamonds$volume = volume
volume_subset = subset(diamonds, diamonds$volume > 0 & diamonds$volume <=800)
cor.test(volume_subset$volume, volume_subset$price)
```


```{r adjustments price vs. volume}
ggplot(aes(x = volume, y = price), data = volume_subset) +
  geom_point(color = 'green') +
  geom_smooth()
```

```{r mean price by clarity}
# Use the function dplyr package
# to create a new data frame containing
# info on diamonds by clarity.

# Name the data frame diamondsByClarity

# The data frame should contain the following
# variables in this order.

#       (1) mean_price
#       (2) median_price
#       (3) min_price
#       (4) max_price
#       (5) n

# where n is the number of diamonds in each
# level of clarity.

diamondsByClarity <- diamonds %>%
  group_by(clarity) %>%
  summarise(mean_price = mean(price),
            median_price = median(as.numeric(price)),#necessary due to bug in version 0.4.1 yesterday (2015-01-14)
            min_price = min(price),
            max_price = max(price),
            n = n()) %>%
  arrange(clarity)

diamondsByClarity

```

```{r two bar plots}
#create data sets
diamonds_by_clarity <- group_by(diamonds, clarity)
diamonds_mp_by_clarity <- summarise(diamonds_by_clarity, mean_price = mean(price))

diamonds_by_color <- group_by(diamonds, color)
diamonds_mp_by_color <- summarise(diamonds_by_color, mean_price = mean(price))

#create plots
p1 = ggplot(aes(x = clarity, y = price), data = diamonds) +
  geom_bar(stat = 'identity', aes(fill = clarity))

p2 = ggplot(aes(x = color, y = price), data = diamonds) +
  geom_bar(stat = 'identity', aes(fill = color))

grid.arrange(p1,p2,ncol = 1)
```
###Observations###
There does not appear to be a strong linear correlation between color or clarity and price.  This leads me to believe there are multiple factors which determine price.  The title of the next set of lessons is a strong indicator of this as well :).

**from course answer**

Mean price tends to decrease as clarity improves. The same can be said for color.

We encourage you to look into the mean price across cut.

```{r mean price as a function of cut}
diamonds_by_cut <- group_by(diamonds, cut)
diamonds_mp_by_cut <- summarise(diamonds_by_cut, mean_price = mean(price))

p3 = ggplot(aes(x = cut, y = price), data = diamonds) +
  geom_bar(stat = 'identity', aes(fill = cut))

grid.arrange(p1,p2,p3,ncol = 1)
```
The relationship between price and cut seems to be linearly correlated.

##Gapminder revisited##

In your investigation, examine pairs of variable and create 2-5 plots that make use of the techniques from Lesson 4.

```{r gapminder revisted}
#load the data set

 workhours <- read.csv("~/Udacity/dataAnalysisR/EDA_R_Udacity/lesson3/workinghours.csv")
names(workhours)

```
###Analyze the data set##

- Start with scatterplot of working hours vs. country and working hours vs. year
- Merge with HPI data set (happiness index for 135 countries)
- Determine if there is a correlation between workhours and happiness

```{r initial gapminder working hours scatterplots}
gmp1 = ggplot(aes( x = year, y = value), data = workhours) +
  geom_point(fill = 'orange')

gmp2 = ggplot(aes( x = country, y = value), data = workhours) +
  geom_point(fill = 'red')

grid.arrange(gmp1,gmp2,ncol =1)
```
This data set is interesting, there are clearly some things to be learned from it, especially near the lower and upper bounds.  To really do something with this data though, we need to add something.  I'm going to merge this with another data set, which contains information such as happiness, gdp, and population by country.

I converted the HPI data set to csv for import.

###Import HPI Dataset###

```{r import and clean HPI Dataset}
hpi <- read.csv(file = '~/Udacity/dataAnalysisR/EDA_R_Udacity/lesson4/hpi.csv')
names(hpi) = tolower(names(hpi))
names(hpi)
```

###Summarise the workhours data###
```{r summarise workhours data frame}
workhourssum <- workhours %>%
  group_by(country) %>%
  summarise(mean_hours = mean(value),
            median_hours = median(as.numeric(value)),#necessary due to bug in version 0.4.1  (2015-01-14)
            min_hours = min(value),
            max_hours = max(value),
            n = n()) %>%
  arrange(max_hours)

head(workhourssum)
tail(workhourssum)
```

###Merge workhours and hpi###
```{r merge workhours and hpi}
merged = merge(workhourssum,hpi, by = 'country')
```

```{r write merged workhours and hpi file to csv}
write.csv(merged, '~/Udacity/dataAnalysisR/EDA_R_Udacity/lesson4/workhours_hpi.csv')

#note that we lost hong kong, korea, czech republic etc. because names did not match, this should be corrected at some point
```

###Plot workhours vs.hpi rank###

```{r workhours vs. hpi plots}
ggplot(aes(x = mean_hours, y = hpi.rank), data = merged) +
  geom_point() +
  geom_smooth(method = 'lm',color = 'red')
```

No surprises here, it looks like there is a negative relationship between hours worked and hpi.rank.

```{r test correlation hpi and hours worked}
cor.test(merged$mean_hours, merged$hpi.rank)
```
There is a weak negative correlation between hours worked and happiness rank.  The gapminder data sest (hours worked) only contained 46 countries (merge key) with the 135 countries in the HPI data set.  This analysis is in sore need for more observations.
