---
title: "Udacity EDA R Final Project"
author: "Andrew Thomas Bauman"
date: "Tuesday, February 10, 2015"
output: html_document
---

#Introduction#

Recent news media coverage and my own personal interest inspired me to explore a school vaccination data set for WA state.  WA DOH employees generously assembled 3 data sets for my project for school years 2011-2012, 2012-2013, and 2013-2014.  The 2011-2012 set is the most comprehensive in terms of school covered and relevant fields  Each data set contains an observation about a school.  The 3 data sets will be concantonated into one data set and then explored.  

Additonal information can be provided by US Census Bureau data and data from the Office of the Superintendent of Public Instruction (OSPI) in Washington State.  This includes geographic data, demograhic data, and economic data at multiple levels of aggregation.  Final data frames will likely include:

- school vaccination data frame for 3 school years (2011-2012,2012-2013,2013-2014)
- Demographic data frame at the school, district, or county level
- school profile data frame
- economic data frame at the zip code level

I will use the following approach to assemble my initial data frame and explore the data:

* Read and audit each of the 3 school vaccination data frames
* Choose columns for the final school vaccination data frame  
* Prepare the 3 data frames to match item 2
* Concatonate the data
* Tidy the data
* Explore the data
* Merge with relevant demographic, economic, or profile data
* Explore the merged data

##Prepare the R Environment##

###Set Path and Load Libraries###

```{r setpath}
setwd("~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject")
```

```{r load libraries}
library(RSocrata) #to access socrata data
library(ggplot2)
library(gridExtra)
library(dplyr)
library(alr3)
library(GGally)
library(reshape2)
library(acs) #to access census data
library(tidyr) #bare bones data tidying library

```
##Read and Audit WA DOH K-12 School Vaccination Data Frames##

###Read WA DOH Data Sets###

Named data sets based on starting year of school year.  That is also how the year variable was assigned.  So 2011 referes to the school year (SY) 2011-2012

```{r load school vaccination exemption data sets}

#WA DOH SY 2011-2012 K-12
doh_k12_2011<- read.csv('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/WA_Vacc_K12_2011.csv')

#WA DOH SY 2012-2013 K-12
doh_k12_2012<- read.csv('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/WA_Vacc_K12_2012.csv')

#WA doh SY 2013-2014 K-12
doh_k12_2013<- read.csv('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/WA_Vacc_K12_2013.csv')
```
###Data Audit###
Most of the audit was done through simple visual inspection of fields and the first 20 rows of data. The columns of the dataframes don't match, there is missing information, and information was recorded in different ways.  The data is not tidy.  Tidying will be saved for the final school vaccination DF.

**Columns for the final school vaccination data frame**
I'm saving some of the geo information, combining some fields, and dropping calculated fields.

year,school_code,school_name,school_type,school_city,school_state,school_zip,school_county,school_district,district_code,grade_levels,reported,enrolled,exempt,medical_exempt,personal_exempt,religious_exempt,exempt_DTTd,exempt_pertussis,exempt_polio,exempt_MMR,exempt_hepB,exempt_varicella

###Prepare Initial Data Frames###
Note that cleaning and auditing is clunky for now while I get acquainted. I may pipe/chain the process, if warranted, at a later date.

####2011####

```{r rename and select columns 2011}

#get columns names
colnames(doh_k12_2011)

#rename
colnames_2011 = c('year','school_code','school_name','school_type','school_street','school_city','school_state','school_zip','school_county','school_district','district_code','grade_levels','reported','enrolled','exempt','percent_exempt','medical_exempt','percent_medical_exempt','personal_exempt','percent_personal_exempt','religious_exempt','percent_religious_exempt','exempt_DTTd','exempt_pertussis','exempt_polio','exempt_measles','exempt_mumps','exempt_rubella','exempt_hepB','exempt_varicella')

colnames(doh_k12_2011) <- colnames_2011

select_colnames_2011 = c('year','school_code','school_name','school_type','school_city','school_state','school_zip','school_county','school_district','district_code','grade_levels','reported','enrolled','exempt','medical_exempt','personal_exempt','religious_exempt','exempt_DTTd','exempt_pertussis','exempt_polio','exempt_measles','exempt_mumps','exempt_rubella','exempt_hepB','exempt_varicella')

#select columns
doh_k12_2011 <- doh_k12_2011[,select_colnames_2011]
```
The 2011 data set contains a field for each component of MMR.  I suspect these are all the same value since I'm not aware of administering MMR components seperately as a common practice.  Are these the same?  If so, I'm going to combine them into just MMR.

```{r audit MMR 2011}

reported_doh_k12_2011 <- subset(doh_k12_2011, doh_k12_2011$reported == 'Yes')
reported_measles <- reported_doh_k12_2011$exempt_measles
reported_mumps <- reported_doh_k12_2011$exempt_mumps
reported_rubella <- reported_doh_k12_2011$exempt_rubella

summary(reported_measles == reported_mumps)
summary(abs(reported_measles - reported_mumps))
maxdiff = reported_doh_k12_2011[max(abs(reported_measles-reported_mumps)),]
```

I don't understand some of the values the summaries are giving Followup inspection is showing very little difference.  It looks like in most or all instances that M, M, and R are the same.  I'm going to use measles as MMR.  I will renamed the column and drop the other two. This will be revisited later.

```{r 2011 drop and rename}
#drop mumps and rubella
doh_k12_2011$exempt_mumps <- NULL
doh_k12_2011$exempt_rubella <- NULL

#rename measles to MMR
names(doh_k12_2011)[names(doh_k12_2011) == 'exempt_measles'] <- 'exempt_MMR'
```

####2013####
```{r prep 2013}
#get columns names
colnames(doh_k12_2013)

#add year, reported, district_code: note initial values except year are placeholders
doh_k12_2013$year <- 2013
doh_k12_2013$reported <- "Yes"
doh_k12_2013$district_code <- 8675309

#rename
colnames_2013 = c('school_code','school_name','school_type','school_street','school_city','school_state','school_zip','school_district','school_county','grade_levels','enrolled',"complete",'exempt','medical_exempt','personal_exempt','religious_exempt','religiousmember_exempt','exempt_DTTd','exempt_pertussis','exempt_polio','exempt_MMR','exempt_hepB','exempt_varicella','year', 'reported',"district_code")

colnames(doh_k12_2013) <- colnames_2013

select_colnames_2013 = c('year','school_code','school_name','school_type','school_city','school_state','school_zip','school_county','school_district','district_code','grade_levels','reported','enrolled','exempt','medical_exempt','personal_exempt','religious_exempt',"religiousmember_exempt",'exempt_DTTd','exempt_pertussis','exempt_polio','exempt_MMR','exempt_hepB','exempt_varicella')

#select columns
doh_k12_2013 <- doh_k12_2013[,select_colnames_2013]


```
#####Handling Religious Exemption#####
It appears that religious exemptions are broken into two categories, religious exemption and religious member exemption.  I will combine these two categories into religious exempt.

```{r 2013 combine religious}
#combine religious exemptions
rel_exempt_2013 <- subset(doh_k12_2013, !is.na(doh_k12_2013$religious_exempt) & !is.na(doh_k12_2013$religiousmember_exempt))

#going to stop here, here is the strategy
#change na values in the relexempt and relmemberexempt columns to a ridiculously large number, probably with a lambda function
#add the columns onto relexempt
#replace anything >= the ridiclously large number with na

doh_k12_2013$religious_exempt[is.na(doh_k12_2013$religious_exempt)]



<- rel_exempt_2013$religious_exempt + rel_exempt_2013$religiousmember_exempt



####2012####



**Important Question**

Are there schools unique to 2012 or 2013?  I need to inspect this to determine if they are worth holding on to and getting geo info.

####School Code Audit####
Going to try subset on this.  I will subset 2012 and 2013 on those school codes != to 2011, if any.  Using the %in% operator and subsetting on False (not in 2011). 
```{r School Code Audit}
#2012 Unique School Code relative to 2011
schoolcode_2012_unique_2011<- subset(doh_k12_2012, doh_k12_2012$School.Code %in% doh_k12_2011$School.Code == F)

  
#2013 Unique School Code relative to 2011
schoolcode_2013_unique_2011<- subset(doh_k12_2013, doh_k12_2013$School.Code %in% doh_k12_2011$School.Code == F)

#2011 unique school Code relative to 2012 and 2013
schoolcode_2011_unique<- subset(doh_k12_2011, (doh_k12_2011$School.Code %in% doh_k12_2012$School.Code == F) & (doh_k12_2011$School.Code %in% doh_k12_2013$School.Code == F))


#2012 unique school code relative to 2011 and 2013
schoolcode_2012_unique<- subset(doh_k12_2012, (doh_k12_2012$School.Code %in% doh_k12_2011$School.Code == F) & (doh_k12_2012$School.Code %in% doh_k12_2013$School.Code == F))

#2013 unique school code relative to 2011 and 2012
schoolcode_2013_unique<- subset(doh_k12_2013, (doh_k12_2013$School.Code %in% doh_k12_2012$School.Code == F) & (doh_k12_2013$School.Code %in% doh_k12_2011$School.Code == F))

#How many of the unique schools are in King County?
summary(schoolcode_2012_unique$County == 'KING')
summary(schoolcode_2013_unique$County == 'KING')
summary(schoolcode_2011_unique$School.County == 'KING')

```
The 2012 data set is missing address data, but there are only 31 unique school codes in this data set, so there are only 31 missing addresses.

Going to rename the 2011 columns.



```{r rename columns}
colnames_2011 = c('year','school_code','school_name','school_type','school_street','school_city','school_state','school_zip','school_county','school_district','district_code','grade_levels','reported','enrolled','exempt','percent_exempt','medical_exempt','percent_medical_exempt','personal_exempt','percent_personal_exempt','religious_exempt','percent_religious_exempt','exempt_DTTd','exempt_pertussis','exempt_polio','exempt_measles','exempt_mumps','exempt_rubella','exempt_hepB','exempt_varicella')

colnames(doh_k12_2011) <- colnames_2011

colnames_2012 = c('school_code','school_name','school_type','school_district','school_county','reported','grade_levels','enrolled','exempt','medical_exempt','personal_exempt','religious_exempt','religious_member_exempt','exempt_DTTd','exempt_pertussis','exempt_polio','exempt_MMR','exempt_hepB','exempt_varicella','year')

colnames(doh_k12_2012) <- colnames_2012


```


```{r 2012 combine religious}
#combine religious exemptions
doh_k12_2011$religious_exempt[]
```

```{r align data to 2011 part 2}

```

