Exploration of WA DOH School Exemption Data over 3 School Years by Andrew T. Bauman
========================================================
#Introduction#

Recent news media coverage and my own personal interest inspired me to explore a school vaccination data set for WA state.  WA DOH employees generously assembled 3 data sets for my project for school years 2011-2012, 2012-2013, and 2013-2014.  The 2011-2012 set is the most comprehensive in terms of school covered and relevant fields  Each data set contains an observation about a school.  The 3 data sets were bound into one data set, then explored.  

The initial data and cleaning/assembly of the combined data frame can be found here:
[githubrepo](https://github.com/baumanab/EDA_R_Udacity/tree/master/EDA_R_FinalProject/EDA_R_FinalProject.Rmd)

Additional cleaning and wrangling can be found here:
[githubrepo](https://github.com/baumanab/EDA_R_Udacity/blob/master/EDA_R_FinalProject/EDA_R_FinalProject_WADOH_DataFrameWrangle.Rmd)

An initial exploration wherein I make decisions about subsetting data for this RMD is and some additoinal corrections for student enrollment, is here:
[githubrepo](https://github.com/baumanab/EDA_R_Udacity/blob/master/EDA_R_FinalProject/WADOH_DataExplore.Rmd)


The current state is a rough approach and far from perfect.  I plan to revisit this as I discover additional elements that need cleaning.  I am currently working on a more readable and organized version of the data wrangling which involves piping/chaining the code together, so that changes are easy to make and push to the .rds file called as data for this project.

**Disclaimer:**
While I am a trained scientist, I am not an expert statistician or programmer.  This is a student project.  My data cleaning and wrangling has room for improvement.  As such, please exercise caution and good judgement regarding my observations and conclusions.  **Feedback on any aspect of this project is encouraged and most welcome.**

There are some aspects of vaccination and the spread of disease through populations that are best left to subject matter experts (SMEs).  I consulted CDC publications throughout this analysis and I encourage the reader to do the same.
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load libraries

library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(GGally)
library(reshape2)
library(tidyr)
library(RColorBrewer)
library(scales)
cb.palette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
                    "#0072B2", "#D55E00", "#CC79A7")

```

```{r echo=FALSE, Load_the_Data}

#school vaccine exemption data, reported == "Yes" for WA State, school years 2011-2012, 2012-2013, 2013-2014
vaccReported <- readRDS("data/vaccReported")


#temp convenience
#setwd("Udacity//dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/")
#save.image("project_pause.RData")
#load("project_pause.RData")
```
##Alter Data##

* Select fields from WA DOH 2011 - 2012, 2012-2013, and 2013 - 2014 School Year data set.*
* Subset or reported only*
* Subset by enrollement 10 - 2500 students all school years*
* Subset by enrollment 10 - 2500 students most recent school year (2013)*
* Subset by enrollment 100 - 2500 students*
``` {r echo = FALSE ,alter_data}
#limit fields
fields <- c("school_year", "school_code", "school_name", "school_type", "school_city","school_state", "school_zip", "school_county", "school_district", "district_code", "enrolled", "total_exemptions", "medical_exempt", "nonmedical_exempt", "exempt_MMR", "percent_exempt", "percent_medical_exempt","percent_nonmedexempt", "percent_exempt_MMR")


vaccReported <- subset(arrange(vaccReported, desc(enrolled)), na.rm = TRUE, select = fields 
                       )
vaccReported_NLT_10_NGT_2500 <- subset(vaccReported,vaccReported$enrolled >= 10 & vaccReported$enrolled <= 2500) %>%
  arrange(desc(percent_exempt))

vaccReported_NLT_10_NGT_2500_2013 <- subset(vaccReported_NLT_10_NGT_2500, school_year == "2013")

vaccReported_NLT_100_NGT_2500 <- subset(vaccReported,vaccReported$enrolled >= 100 & vaccReported$enrolled <= 2500) %>%
  arrange(desc(percent_exempt))

```

# Univariate Plots Section

##Distribution of enrolled students##
```{r echo=FALSE, dist_enrollment}
#vacc
names(vaccReported)
str(vaccReported)
summary(vaccReported)

#distribution of enrollment
ggplot(aes(x = enrolled), data = vaccReported) + 
  geom_histogram(binwidth = 30,  color = 'black',fill = '#F79420') +
  scale_x_continuous(limits = c(0,5100), breaks = seq(0, 5100, 500))
```
This is a long tailed distribution, a feature exhibited by most distributions in this data set.  Log transformations are performed to visualize the data.
```{r echo = False dist_enrollment_cont}



brks <- c(1, 5, 10, 50, 100, 500, 1000, 5000, 10000)

ggplot(aes(x = enrolled), data = vaccReported) + 
  geom_histogram(binwidth = log(1.25),  color = 'black',fill = '#F79420') +
  scale_x_log10(breaks = brks) +
  xlab("Log10(enrolled)")

```
Range:  1 - 5076
Most of the data falls between ~ 100 - 1000 students per school

##Distribution of total exemptions##
```{r echo=FALSE, total_exemptions}


#summary of total exemptions
summary(vaccReported$total_exemptions)
by(vaccReported$total_exemptions, vaccReported$school_year,summary)

#total exemption rate
summary(vaccReported$percent_exempt)
by(vaccReported$percent_exempt, vaccReported$school_year,summary)


#distribution of total exemptions

ggplot(aes(x = total_exemptions), data = vaccReported) + 
  geom_histogram(binwidth = 5,  color = 'black',fill = '#F79420') +
  scale_x_continuous(limits = c(0,450), breaks = seq(0, 450, 50))


ggplot(aes(x = total_exemptions), data = vaccReported) + 
  geom_histogram(binwidth = log(1.2),  color = 'black',fill = '#F79420') +
  scale_x_log10(breaks = brks) +
  xlab("Log10(Total Exemptions)")
  

#total exempt by year

ggplot(aes(x = total_exemptions), data = vaccReported) + 
  geom_histogram(binwidth =log(1.25),  color = 'black',fill = '#F79420') +
  facet_wrap(~school_year) +
  scale_x_log10(breaks = brks) +
  xlab("Log10(Total Exemptions)")



#percent total exemptions
brks2 <- c(.25,.5, 1, 2.5, 5, 10, 15, 25, 50, 75, 100)

ggplot(aes(x = percent_exempt, y = ..density..), data = vaccReported) +
  geom_freqpoly(binwidth = .1, aes(color = school_year)) +
  scale_x_log10(breaks = brks2)
  
##freqpoly proportion of percent exempt, by county, color = year
ggplot(aes(x = percent_exempt, y = ..count../sum(..count..)), data = vaccReported) +
  geom_freqpoly(binwidth = log(1.25), aes(color = school_year)) +
  facet_wrap(~school_county) +
  scale_x_log10()



##freqpoly density of percent exempt, by county, color = year
ggplot(aes(x = percent_exempt, y = ..density..), data = vaccReported) +
  geom_freqpoly(binwidth = .5, aes(color = school_year)) +
  facet_wrap(~school_county) +
  scale_x_log10()

#plot select counties to compare to King
  
ggplot(aes(x = percent_exempt, y = ..density..), data = subset(vaccReported, school_county == "KING" | school_county == "FERRY" | school_county == "CLARK" | school_county == "PIERCE" | school_county == "SAN JUAN")) +
  geom_freqpoly(binwidth = .5, aes(color = school_county, size = 1.25, alpha = .5)) +
  scale_x_log10()

```
* Range: 0 - 436*
* Rate:  0 - 100%, ~ 6.6% mean for all 3 school years*
* The highest density data is from ~ 10 - 75 students per school*
* Year to year distributions are similiar for total and rate*
* most schools fall into a rate range of 2-25%*
* King County has the highest proportion of exemptions*
* There appears to be some smaller counties (i.e. Ferry, San Juan) and others with more density at higher exemption rates than King and Pierce county (most populous counties in WA)*


**Note:**  Higher exemption rates of ~ > 50% appear to be associated with lower levels of enrollment (< 100 students).  This will be investigated in the bivariate section.




```{r echo=FALSE, medical_exemptions}



#explore medical exemptions
summary(vaccReported$medical_exempt)
by(vaccReported$medical_exempt, vaccReported$school_year,summary)

summary(vaccReported$percent_medical_exempt)
by(vaccReported$percent_medical_exempt, vaccReported$school_year,summary)

#medical exempt

ggplot(aes(x = medical_exempt), data = vaccReported) + 
  geom_histogram(binwidth = log(1.25),  color = 'black',fill = '#F79420') +
  scale_x_log10(breaks = brks) +
  xlab("Log10(medical_exemptions)")

ggplot(aes(x = medical_exempt), data = vaccReported) + 
  geom_histogram(binwidth = log(1.25),  color = 'black',fill = '#F79420') +
  facet_wrap(~school_year) +
  scale_x_log10(breaks = brks) +
  xlab("Log10(medical_exemptions)")

#percent medical exemptions


ggplot(aes(x = percent_medical_exempt, y = ..density..), data = vaccReported) +
  geom_freqpoly(binwidth = .1, aes(color = school_year)) +
  scale_x_log10()
  
##freqpoly proportion of percent medical exempt, by county, color = year
ggplot(aes(x = percent_medical_exempt, y = ..count../sum(..count..)), data = subset(vaccReported, vaccReported$percent_medical_exempt > 0)) +
  geom_freqpoly(binwidth = .5, aes(color = school_year)) +
  facet_wrap(~school_county) +
  scale_x_log10()

##freqpoly density of percent medical exempt, by county, color = year
ggplot(aes(x = percent_medical_exempt, y = ..density..), data = vaccReported) +
  geom_freqpoly(binwidth = .5, aes(color = school_year)) +
  facet_wrap(~school_county) +
  scale_x_log10()

#explore non-medical exemptions
summary(vaccReported$nonmedical_exempt)
by(vaccReported$nonmedical_exempt, vaccReported$school_year,summary)

summary(vaccReported$percent_nonmedexempt)
by(vaccReported$percent_nonmedexempt, vaccReported$school_year,summary)


```

```{r echo=FALSE, nonmedical_exemptions}


#non-medical exempt



ggplot(aes(x = nonmedical_exempt), data = vaccReported) + 
  geom_histogram(binwidth = .1,  color = 'black',fill = '#F79420') +
  scale_x_log10() +
  xlab("Log10(nonmedical_exemptions)")

ggplot(aes(x = nonmedical_exempt), data = vaccReported) + 
  geom_histogram(binwidth = .1,  color = 'black',fill = '#F79420') +
  facet_wrap(~school_year) +
  scale_x_log10() +
  xlab("Log10(nonmedical_exemptions)")


#percent non-medical exemptions


ggplot(aes(x = percent_nonmedexempt, y = ..density..), data = vaccReported) +
  geom_freqpoly(binwidth = .1, aes(color = school_year)) +
  scale_x_log10()
  
##freqpoly proportion of percent nonmedical exempt, by county, color = year
ggplot(aes(x = percent_nonmedexempt, y = ..count../sum(..count..)), data = vaccReported) +
  geom_freqpoly(binwidth = .5, aes(color = school_year)) +
  facet_wrap(~school_county) +
  scale_x_log10()

##freqpoly density of percent nonmedical exempt, by county, color = year
ggplot(aes(x = percent_nonmedexempt, y = ..density..), data = vaccReported) +
  geom_freqpoly(binwidth = .5, aes(color = school_year)) +
  facet_wrap(~school_county) +
  scale_x_log10()

```

```{r echo=FALSE, MMR_exemptions}


#MMR exemptions

summary(vaccReported$exempt_MMR)
by(vaccReported$exempt_MMR, vaccReported$school_year,summary)

summary(vaccReported$percent_exempt_MMR)
by(vaccReported$percent_exempt_MMR, vaccReported$school_year,summary)




ggplot(aes(x = exempt_MMR), data = vaccReported) + 
  geom_histogram(binwidth = .1,  color = 'black',fill = '#F79420') +
  scale_x_log10() 


ggplot(aes(x = exempt_MMR), data = vaccReported) + 
  geom_histogram(binwidth = .1,  color = 'black',fill = '#F79420') +
  facet_wrap(~school_year) +
  scale_x_log10() 
  

#percent exempt MMR


ggplot(aes(x = percent_exempt_MMR, y = ..density..), data = vaccReported) +
  geom_freqpoly(binwidth = .1, aes(color = school_year)) +
  scale_x_log10()
  
##freqpoly proportion of percent exempt MMR, by county, color = year
ggplot(aes(x = percent_exempt_MMR, y = ..count../sum(..count..)), data = vaccReported) +
  geom_freqpoly(binwidth = .5, aes(color = school_year)) +
  facet_wrap(~school_county) +
  scale_x_log10()

##freqpoly density of percent exempt MMR, by county, color = year
ggplot(aes(x = percent_exempt_MMR, y = ..density..), data = vaccReported) +
  geom_freqpoly(binwidth = .5, aes(color = school_year)) +
  facet_wrap(~school_county) +
  scale_x_log10()

ggplot(aes(x = percent_exempt_MMR, y = ..density..), data = vaccReported) +
  geom_freqpoly(binwidth = .5, aes(color = school_year)) +
  facet_wrap(~school_county) +
  scale_x_log10()



```

```{r echo=FALSE, exemption_types}


melt <- gather(vaccReported_NLT_10_NGT_2500,exemption_type,exemptions,medical_exempt, nonmedical_exempt)

#add percent by exemption type
melt$percent_exemption_type <- round((melt$exemptions/melt$total_exemptions)*100,digits = 2)

ggplot(data = melt, aes(x = percent_exemption_type, y = ..density..)) +
  geom_freqpoly(binwidth = log(1.25),aes(color = exemption_type)) +
  facet_wrap(school_type ~ school_year) +
  scale_x_log10()

ggplot(data = melt, aes(x = percent_exemption_type, y = ..count../sum(..count..))) +
  geom_freqpoly(binwidth = log(1.25),aes(color = exemption_type)) +
  facet_wrap(school_type ~ school_year) +
  scale_x_log10()


ggplot(data = subset(melt, exemption_type == 'medical_exempt'), aes(x = percent_exemption_type, y = ..density..)) +
  geom_freqpoly(binwidth = log(1.25),aes(color = school_type)) +
  facet_wrap(~school_year) +
  scale_x_log10()

ggplot(data = subset(melt, exemption_type == 'nonmedical_exempt'), aes(x = percent_exemption_type, y = ..density..)) +
  geom_freqpoly(binwidth = log(1.25),aes(color = school_type)) +
  facet_wrap(~school_year) +
  scale_x_log10()

ggplot(data = subset(melt, exemption_type == 'nonmedical_exempt'), aes(x = percent_exemption_type, y = ..count../sum(..count..))) +
  geom_freqpoly(binwidth = log(1.1),aes(color = school_year)) +
  scale_x_log10()

ggplot(data = subset(melt, exemption_type == 'nonmedical_exempt'), aes(x = percent_exemption_type, y = ..count../sum(..count..))) +
  geom_freqpoly(binwidth = log(1.25),aes(color = school_type)) +
  facet_wrap(~school_year) +
  scale_x_log10()
```

```{r echo=FALSE, group_by_attribute}


#change nonmedical exempt to 20 and then resave .rds
inspect <- vaccReported[vaccReported$school_name == "Tacoma Waldorf School",]

#going to do for 2013 (most recent year) top 10 by mean nonmedical exempt for county, then withing that county, district, then city, then withing city the zip code.  Then get count of schools in each that are over 10% total exempt, and 5% MMR exempt.  Maybe save some of this for bivariate.  I'm thinking that in bivariate I can look at schools or zones over the MMR threshold and then determine what proportion of the total exemptions are from medical and which are from nonmedical in those zones.  that could be a final plot, the point being is taht choice is pushing it over.

by_county_exempt <- vaccReported_NLT_10_NGT_2500_2013 %>%  group_by(school_county) %>% 
  summarize(mean_percent_exempt = round(mean(percent_exempt, na.rm=TRUE), 2), 
            median_percent_exempt = round(median(percent_exempt, na.rm=TRUE), 2), 
            max_percent_exempt = round(max(percent_exempt, na.rm=TRUE), 2), 
            min_percent_exempt = round(min(percent_exempt, na.rm=TRUE), 2), 
            schools=n(), 
            students=sum(enrolled, na.rm=TRUE)) %>% 
  na.omit() %>% 
  arrange(desc(mean_percent_exempt))

by_county_nonmedexempt <- vaccReported_NLT_10_NGT_2500_2013 %>%  group_by(school_county) %>% 
  summarize(mean_percent_nonmedexempt = round(mean(percent_nonmedexempt, na.rm=TRUE), 2), 
            median_percent_nonmedexempt = round(median(percent_nonmedexempt, na.rm=TRUE), 2), 
            max_percent_nonmedexempt = round(max(percent_nonmedexempt, na.rm=TRUE), 2), 
            min_percent_nonmedexempt = round(min(percent_nonmedexempt, na.rm=TRUE), 2), 
            schools=n(), 
            students=sum(enrolled, na.rm=TRUE)) %>% 
  na.omit() %>% 
  arrange(desc(mean_percent_nonmedexempt))


by_city_exempt <- vaccReported_NLT_10_NGT_2500_2013 %>%  group_by(school_city) %>% 
  summarize(mean_percent_exempt = round(mean(percent_exempt, na.rm=TRUE), 2), 
            median_percent_exempt = round(median(percent_exempt, na.rm=TRUE), 2), 
            max_percent_exempt = round(max(percent_exempt, na.rm=TRUE), 2), 
            min_percent_exempt = round(min(percent_exempt, na.rm=TRUE), 2), 
            schools=n(), 
            students=sum(enrolled, na.rm=TRUE)) %>% 
  na.omit() %>% 
  arrange(desc(mean_percent_exempt))

by_city_nonmedexempt <- vaccReported_NLT_10_NGT_2500_2013 %>%  group_by(school_city) %>% 
  summarize(mean_percent_nonmedexempt = round(mean(percent_nonmedexempt, na.rm=TRUE), 2), 
            median_percent_nonmedexempt = round(median(percent_nonmedexempt, na.rm=TRUE), 2), 
            max_percent_nonmedexempt = round(max(percent_nonmedexempt, na.rm=TRUE), 2), 
            min_percent_nonmedexempt = round(min(percent_nonmedexempt, na.rm=TRUE), 2), 
            schools=n(), 
            students=sum(enrolled, na.rm=TRUE)) %>% 
  na.omit() %>% 
  arrange(desc(mean_percent_nonmedexempt))


by_zip_exempt <- vaccReported_NLT_10_NGT_2500_2013 %>%  group_by(school_zip) %>% 
  summarize(mean_percent_exempt = round(mean(percent_exempt, na.rm=TRUE), 2), 
            median_percent_exempt = round(median(percent_exempt, na.rm=TRUE), 2), 
            max_percent_exempt = round(max(percent_exempt, na.rm=TRUE), 2), 
            min_percent_exempt = round(min(percent_exempt, na.rm=TRUE), 2), 
            schools=n(), 
            students=sum(enrolled, na.rm=TRUE)) %>% 
  na.omit() %>% 
  arrange(desc(mean_percent_exempt))

by_zip_nonmedexempt <- vaccReported_NLT_10_NGT_2500_2013 %>%  group_by(school_zip) %>% 
  summarize(mean_percent_nonmedexempt = round(mean(percent_nonmedexempt, na.rm=TRUE), 2), 
            median_percent_nonmedexempt = round(median(percent_nonmedexempt, na.rm=TRUE), 2), 
            max_percent_nonmedexempt = round(max(percent_nonmedexempt, na.rm=TRUE), 2), 
            min_percent_nonmedexempt = round(min(percent_nonmedexempt, na.rm=TRUE), 2), 
            schools=n(), 
            students=sum(enrolled, na.rm=TRUE)) %>% 
  na.omit() %>% 
  filter(students >=500) %>%
  arrange(desc(mean_percent_nonmedexempt))
  



```
```{r temp sandbox}






### by specific type
by.mwc <- data.sub %>%  group_by(MWC) %>% summarize(Mean.PBE = round(mean(Exempt, na.rm=TRUE), 2), Median.PBE = round(median(Exempt, na.rm=TRUE), 2), Max.PBE = round(max(Exempt, na.rm=TRUE), 2), Min.PBE = round(min(Exempt, na.rm=TRUE), 2), Schools=n(), Students=sum(enrollment, na.rm=TRUE)) %>% na.omit() %>% arrange(desc(Mean.PBE))



state_rate <- mean(vaccReported$percent_exempt, na.rm=T)
#state_rate_NLT100 <- mean(vaccReported_NLT100$percent_exempt, na.rm = T)

#summarize(vaccReported_NLT100, Mean_percent_exempt = mean(percent_exempt, na.rm = TRUE))

#summarize(vaccReported_NLT100, Mean_percent_nonmedexempt = mean(percent_nonmedexempt, na.rm = TRUE))

#boxplot to compare means/medians NLT100 vs reg



##total percent exempt all schools

##state level by year
ggplot(vaccReported, aes(x = school_state, y = percent_exempt)) +
  geom_boxplot() +
  geom_jitter(alpha = .5, aes(size = enrolled, color = enrolled)) +
  facet_wrap(~school_year) +
  scale_color_gradientn(colours = c("red", "orange", "yellow")) #fix color scheme and do plots where outier pops are removed

#year level by county
ggplot(vaccReported, aes(x = school_year, y = percent_exempt)) +
  geom_boxplot(outlier.colour = "red") +
  facet_wrap(~school_county)

##total percent exempt all schools

##state level by year
#ggplot(vaccReported_NLT100, aes(x = school_state, y = percent_exempt)) +
 # geom_boxplot() +
  #geom_jitter(alpha = .5, aes(size = enrolled, color = enrolled)) +
  #facet_wrap(~school_year) +
  #scale_color_gradientn(colours = c("red", "orange", "yellow")) #fix color scheme and do plots where outier pops are removed

##year level by county
#ggplot(vaccReported_NLT100, aes(x = school_year, y = percent_exempt)) +
 # geom_boxplot(outlier.colour = "red") +
  #facet_wrap(~school_county)






#exemption rate vs enrollment
p = ggplot(vaccReported, aes(x = enrolled, y = percent_exempt)) + 
  geom_point(alpha = .4, aes(color = "red"))  
  
p1 = p +  geom_line(alpha = .5,stat = 'summary',fun.y = mean, linetype = 1, color = "black", size = 1) 

p2 = p1 + facet_wrap(~school_year)

p3 = p1 + geom_vline(xintercept = 2500, color="orange", size = 1) +
     geom_vline(xintercept = 2000, color="orange", size = 1)  
  

p4 = p3 + scale_x_continuous(limits = c(0, 2000), breaks = seq(0,2000,50))

p5 = p4 + geom_vline(xintercept = 100, color="orange", size = 1) +
  geom_vline(xintercept = 25, color="orange", size = 1)
  

p1
p2
p3
p4
p5




```

# Univariate Analysis

### What is the structure of your dataset?
My data set contains 6696 observat

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?



# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

### What was the strongest relationship you found?




# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection


# References
