---
title: "Udacity EDA R Final Project"
author: "Andrew Thomas Bauman"
date: "Tuesday, February 10, 2015"
output: html_document
---

#Introduction#

Recent news media coverage and my own personal interest inspired me to explore a school vaccination data set for WA state.  WA DOH employees generously assembled 3 data sets for my project for school years 2011-2012, 2012-2013, and 2013-2014.  The 2011-2012 set is the most comprehensive in terms of school covered and relavant fields.  I will use the following approach:

- Align 2012 and 2013 data with 2011 data, clean, and homogenize
- Tidy the data
- Concatonate the data
- Merge with demographic data
- Explore the data

###Set Path and Load Libraries###

```{r setpath}
setwd("~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject")
```

```{r load libraries}
library(RSocrata) #to access socrata data
library(ggplot2)
library(gridExtra)
library(dplyr)
library(alr3)
library(GGally)
library(reshape2)
library(acs) #to access census data
library(tidyr) #bare bones data tidying library

```

###Load WA DOH Data Sets###

Named data sets based on starting year of school year.  That is also how the year variable was assigned.  So 2011 referese to the school year (SY) 2011-2012

```{r load school vaccination exemption data sets}

#WA DOH SY 2011-2012 K-12
doh_k12_2011<- read.csv('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/WA_Vacc_K12_2011.csv')

#WA DOH SY 2012-2013 K-12
doh_k12_2012<- read.csv('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/WA_Vacc_K12_2012.csv')

#WA doh SY 2013-2014 K-12
doh_k12_2013<- read.csv('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/WA_Vacc_K12_2013.csv')
```
###Data Audit###
Most of the audit was done through simple visual inspection.  

```{r Data Audit}

```

###Align Columns to 2011 Set###
Initial cleaning etc. may be clunky and manual.  I intend to semi-automate this with a pipeline prior to project completion.

```{r align to 2011}
#add Year
doh_k12_2012$Year = 2012
doh_k12_2013$Year= 2013

#drop Data.Source
doh_k12_2012$Data.Source <- NULL
```


