---
title: "Udacity EDA R Final Project Part 1: Data Wrangling"
author: "Andrew Thomas Bauman"
date: "Tuesday, February 10, 2015"
output: html_document
---

#Introduction#

Recent news media coverage and my own personal interest inspired me to explore a school vaccination data set for WA state.  WA DOH employees generously assembled 3 data sets for my project for school years 2011-2012, 2012-2013, and 2013-2014.  The 2011-2012 set is the most comprehensive in terms of school covered and relevant fields  Each data set contains an observation about a school.  The 3 data sets will be concantonated into one data set and then explored.  

Additonal information can be provided by US Census Bureau data and data from the Office of the Superintendent of Public Instruction (OSPI) in Washington State.  This includes geographic data, demograhic data, and economic data at multiple levels of aggregation.  Final data frames will likely include:

- school vaccination data frame for 3 school years (2011-2012,2012-2013,2013-2014)
- Demographic data frame at the school, district, or county level
- school profile data frame
- economic data frame at the zip code level

I will use the following approach to assemble my initial data frame and explore the data:

* Read and audit each of the 3 school vaccination data frames
* Choose columns for the final school vaccination data frame  
* Prepare the 3 data frames to match item 2
* Concatonate the data
* Tidy the data
* Explore the data
* Merge with relevant demographic, economic, or profile data
* Explore the merged data

##Prepare the R Environment##

###Set Path and Load Libraries###

```{r setpath}
load("~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/project_pause.RData")
setwd("~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject")
```

```{r load libraries}
library(dplyr)
library(alr3)
library(tidyr) #bare bones data tidying library

```
##Read and Audit WA DOH K-12 School Vaccination Data Frames##

###Read WA DOH Data Sets###

Named data sets based on starting year of school year.  That is also how the year variable was assigned.  So 2011 refers to the school year (SY) 2011-2012

Data was provided as 3 Excel files.  The files contained summary functions as well some formatting and images at the bottom of the files.  I didn't realize this at first and created .csv and .txt files for import.  The files imported but many columns contained incorrect data classes (integers as strings).  Removing the formatting, images, and summary functions reduced this considerably.  I have provided the orignal files as a .rar on github should someone want to audit my methods for loss.

```{r load school vaccination exemption data sets}

#WA DOH SY 2011-2012 K-12
doh_k12_2011<- read.delim('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/2011.txt', quote="" , stringsAsFactors=FALSE)

#WA DOH SY 2012-2013 K-12
doh_k12_2012<- read.delim('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/2012.txt', quote="" , stringsAsFactors=FALSE)

#WA doh SY 2013-2014 K-12
doh_k12_2013<- read.delim('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/2013.txt', quote="" , stringsAsFactors=FALSE)

#checking structures to determine state of data and column classes
str(doh_k12_2011)
str(doh_k12_2012)
str(doh_k12_2013)
```
###Data Audit###
Most of the audit was done through simple visual inspection of fields and the first 20 rows of data. The columns of the dataframes don't match, there is missing information, and information was recorded in different ways.  The data is not tidy.  Tidying will be saved for the final school vaccination DF.

**Columns for the final school vaccination data frame**
I'm saving some of the geo information, combining some fields, and dropping calculated fields.

year,school_code,school_name,school_type,school_city,school_state,school_zip,school_county,school_district,district_code,grade_levels,reported,enrolled,exempt,medical_exempt,personal_exempt,religious_exempt,exempt_DTTd,exempt_pertussis,exempt_polio,exempt_MMR,exempt_hepB,exempt_varicella

###Prepare Initial Data Frames###
Note that cleaning and auditing is clunky for now while I get acquainted. I may pipe/chain the process, if warranted, at a later date.

####2011####

```{r rename and select columns 2011}
#get columns names
colnames(doh_k12_2011)

#rename
colnames_2011 = c('year','school_code','school_name','school_type','school_street','school_city','school_state','school_zip','school_county','school_district','district_code','grade_levels','reported','enrolled','exempt','percent_exempt','medical_exempt','percent_medical_exempt','personal_exempt','percent_personal_exempt','religious_exempt','percent_religious_exempt','exempt_DTTd','exempt_pertussis','exempt_polio','exempt_measles','exempt_mumps','exempt_rubella','exempt_hepB','exempt_varicella')

colnames(doh_k12_2011) <- colnames_2011

select_colnames_2011 = c('year','school_code','school_name','school_type','school_city','school_state','school_zip','school_county','school_district','district_code','grade_levels','reported','enrolled','exempt','medical_exempt','personal_exempt','religious_exempt','exempt_DTTd','exempt_pertussis','exempt_polio','exempt_measles','exempt_mumps','exempt_rubella','exempt_hepB','exempt_varicella')

#select columns
doh_k12_2011 <- doh_k12_2011[,select_colnames_2011]
```
The 2011 data set contains a field for each component of MMR.  I suspect these are all the same value since I'm not aware of administering MMR components seperately as a common practice.  Are these the same?  If so, I'm going to combine them into just MMR.

```{r audit MMR 2011}

reported_doh_k12_2011 <- subset(doh_k12_2011, doh_k12_2011$reported == 'Yes')
reported_measles <- reported_doh_k12_2011$exempt_measles
reported_mumps <- reported_doh_k12_2011$exempt_mumps
reported_rubella <- reported_doh_k12_2011$exempt_rubella

summary(reported_measles == reported_mumps)
summary(abs(reported_measles - reported_mumps))
maxdiff = reported_doh_k12_2011[max(abs(reported_measles-reported_mumps)),]
```

I don't understand some of the values the summaries are giving. Followup inspection reveals very little difference.  It looks like in most or all instances that M, M, and R are identical or similiar.  I'm going to use measles as MMR.  I will rename the column and drop the other two. This will be revisited later.

```{r 2011 drop and rename}
#drop mumps and rubella
doh_k12_2011$exempt_mumps <- NULL
doh_k12_2011$exempt_rubella <- NULL

#rename measles to MMR
names(doh_k12_2011)[names(doh_k12_2011) == 'exempt_measles'] <- 'exempt_MMR'
```

####2013####
```{r prep 2013}
#get columns names
colnames(doh_k12_2013)

#add year, reported, district_code: note initial values except year are placeholders
doh_k12_2013$year <- 2013
doh_k12_2013$reported <- "Yes"
doh_k12_2013$district_code <- 8675309

#rename
colnames_2013 = c('school_code','school_name','school_type','school_street','school_city','school_state','school_zip','school_district','school_county','grade_levels','enrolled',"complete",'exempt','medical_exempt','personal_exempt','religious_exempt','religiousmember_exempt','exempt_DTTd','exempt_pertussis','exempt_polio','exempt_MMR','exempt_hepB','exempt_varicella','year', 'reported',"district_code")

colnames(doh_k12_2013) <- colnames_2013

select_colnames_2013 = c('year','school_code','school_name','school_type','school_city','school_state','school_zip','school_county','school_district','district_code','grade_levels','reported','enrolled','exempt','medical_exempt','personal_exempt','religious_exempt',"religiousmember_exempt",'exempt_DTTd','exempt_pertussis','exempt_polio','exempt_MMR','exempt_hepB','exempt_varicella')

#select columns
doh_k12_2013 <- doh_k12_2013[,select_colnames_2013]


```
#####Handling Religious Exemption#####
It appears that religious exemptions are broken into two categories, religious exemption and religious member exemption.  I will combine these two categories into religious exempt.

```{r 2013 combine religious}
#assign is.na to unrealistic number
doh_k12_2013$religious_exempt[is.na(doh_k12_2013$religious_exempt)] <- 5555
doh_k12_2013$religiousmember_exempt[is.na(doh_k12_2013$religiousmember_exempt)] <- 5555

#combine religious exemption
doh_k12_2013$religious_exempt <- doh_k12_2013$religious_exempt + doh_k12_2013$religiousmember_exempt

#drop religiousmember_exempt
doh_k12_2013$religiousmember_exempt <- NULL

#reassign na
doh_k12_2013$religious_exempt[doh_k12_2013$religious_exempt >= 5555] <- NA

```
#####Handling Reported#####
The data set did not originally have a field indicating whether a school reported.
I'm assuming that schools that have a value, even a 0 value, for exemptions, reported, and those with na (original value of file is blank), did not report.  I noticed that for the blank cases in exempt, most have a blank for enrollment also.  

```{r 2013 assgning reported}
doh_k12_2013$reported[is.na(doh_k12_2013$exempt)] <- "No"

```




####2012####
```{r add & rename columns 2012}
#add year
doh_k12_2012$year <- 2012


#rename columns
colnames_2012 = c('school_code','school_name','school_type','school_district','school_county','reported',"data_source",'grade_levels','enrolled',"complete",'exempt','medical_exempt','personal_exempt','religious_exempt','religiousmember_exempt','exempt_DTTd','exempt_pertussis','exempt_polio','exempt_MMR','exempt_hepB','exempt_varicella','year')

colnames(doh_k12_2012) <- colnames_2012

#drop data_source and complete
doh_k12_2012$data_source <- NULL
doh_k12_2012$complete <- NULL



```
#####Handling Religous Exemptions#####
This is the same situation as the 2013 data, with the same solution.
```{r religious exemptions 2012}
#assign is.na to unrealistic number
doh_k12_2012$religious_exempt[is.na(doh_k12_2012$religious_exempt)] <- 5555
doh_k12_2012$religiousmember_exempt[is.na(doh_k12_2012$religiousmember_exempt)] <- 5555

#combine religious exemption
doh_k12_2012$religious_exempt <- doh_k12_2012$religious_exempt + doh_k12_2012$religiousmember_exempt

#drop religiousmember_exempt
doh_k12_2012$religiousmember_exempt <- NULL

#reassign na
doh_k12_2012$religious_exempt[doh_k12_2012$religious_exempt >= 5555] <- NA

```



###Fill in missing zip codes and district codes###
The 2012 data set is missing city, state, zip, and district code.  The 2013 data set is missing district code.  The 2012, 2013, and 2011 data sets have a lot of overlap, and all of them have the school code field.  I will write a function that reads in school code and returns zip code, district code, and zip.

```{r missing values 2012}
#fill in state
doh_k12_2012$state <- "WA"


```
##Fill in missing zip and district codes 2012 & 2013##
**Plan**
- assemble a dataframe of unique school codes, zip codes, school city, and disrict codes from OSPI and current data sets
- Use lambda function to fill in missing codes
###Unique Schools by Year###

Are there schools unique to any year?  This would be reasonable since schools open and close each year in WA.

###Construct school profile keys and fill in gaps###
**Strategy**  I have some options for filling in missing geo data.  I can fill in from the WA DOH data sets, from the OSIP data sets, or both.  Note that the OSIP data sets have complete zip codes, whereas the WA DOH data sets have only the 5 digit zip code.  The complete zip code may come in handy later for extracting census info.  The 5 digit would be good for aggregation on overall zip code.
**NOte on Packages** The ACS package has an apply function.  I used detach(package:acs) to remove the acs package so that R base apply would not be masked.

I'm going to use all the data stats for filing in info and Split, Apply, Combine

- assemble WA DOH and OSIP codeframes (geo info keys including school building code vector)
- apply on OSIP first
- split on missing values
- fill in on WA DOH
- Rbind to original data sets
- create two zip code vectors, 1 for first five and 1 for last digits.  Later I can combine these for acs queries.

```{r constrct WA DOH codeframe keys}
codeframe_2011 <- doh_k12_2011[c("school_code","school_zip","school_city","district_code")]

codeframe_2013 <- doh_k12_2013[c("school_code","school_zip","school_city","district_code")]

WA_DOH_codeframe <- rbind(codeframe_2011,codeframe_2013)

#remove duplicate school codes
WA_DOH_codeframe <- WA_DOH_codeframe[!duplicated(WA_DOH_codeframe$school_code),]

```
```{r Read in OSIP data and construct OSIP codeframes}
#WA OSIP 1
WA_OSIP_1<- read.delim('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/school_lists/482_SDF_bldg.txt', quote="" , stringsAsFactors=FALSE)

#WA OSIP 2
WA_OSIP_2<- read.delim('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/school_lists/483_SDF_bldg.txt', quote="" , stringsAsFactors=FALSE)

#WA OSIP 3
WA_OSIP_3<- read.delim('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/school_lists/491_SDF_ApprovedPrivateSchoolsList.txt', quote="" , stringsAsFactors=FALSE)

#WA OSIP 4
WA_OSIP_4<- read.delim('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/school_lists/492_SDF_ApprovedPrivateSchoolsList.txt', quote="" , stringsAsFactors=FALSE)

names(WA_OSIP_2) == names(WA_OSIP_1)

names(WA_OSIP_3) == names(WA_OSIP_4)


```


####Plan for OSIP Data Frames####
- combine public
- combine private
- audit and clean
- add subset with unique school codes relative to WA DOH Codeframe to WA DOH codeframe

```{r combine OSIP data frames public set and private set}
#rename columns
names(WA_OSIP_1)[3] <- "district_code"
names(WA_OSIP_1)[5] <- "school_code"
names(WA_OSIP_1)[11] <- "school_city"
names(WA_OSIP_1)[13] <- "school_zip"

names(WA_OSIP_2)[3] <- "district_code"
names(WA_OSIP_2)[5] <- "school_code"
names(WA_OSIP_2)[11] <- "school_city"
names(WA_OSIP_2)[13] <- "school_zip"

names(WA_OSIP_3)[4] <- "school_code"
names(WA_OSIP_3)[7] <- "school_city"
names(WA_OSIP_3)[9] <- "school_zip"

names(WA_OSIP_4)[4] <- "school_code"
names(WA_OSIP_4)[7] <- "school_city"
names(WA_OSIP_4)[9] <- "school_zip"

#combine rows

cols_public <- c("school_code", "school_city","school_zip", "district_code")
cols_private <- c("school_code", "school_city","school_zip")

OSIP_codeframe_public<- rbind(WA_OSIP_1[cols_public],WA_OSIP_2[cols_public])

OSIP_codeframe_private<- rbind(WA_OSIP_3[cols_private],WA_OSIP_4[cols_private])

#clean city field (web addresses and empty)
OSIP_codeframe_private <- OSIP_codeframe_private[grepl("www",OSIP_codeframe_private$school_city) == F,]
OSIP_codeframe_private <- OSIP_codeframe_private[grepl(" ",OSIP_codeframe_private$school_city) == F,]

OSIP_codeframe_public <- OSIP_codeframe_public[grepl("www",OSIP_codeframe_public$school_city) == F,]
OSIP_codeframe_public <- OSIP_codeframe_public[grepl(" ",OSIP_codeframe_public$school_city) == F,]

#srip zipcode at hypehn ("-"), extract first five digits
OSIP_codeframe_private$school_zip <- sapply(strsplit(OSIP_codeframe_private$school_zip,split = "-"),"[",1)
OSIP_codeframe_public$school_zip <- sapply(strsplit(OSIP_codeframe_public$school_zip,split = "-"),"[",1)



#omit NA
OSIP_codeframe_public <- na.omit(OSIP_codeframe_public)
OSIP_codeframe_private <- na.omit(OSIP_codeframe_private)

```
```{r assembling final codeframe}
#how many OSIP codeframe private school codes not in WA DOH codeframe?
OSIP_codeframe_private_unique<- subset(OSIP_codeframe_private, OSIP_codeframe_private$school_code %in% WA_DOH_codeframe$school_code == F)

#private unique with actual zip
OSIP_codeframe_private_unique <- OSIP_codeframe_private_unique[grepl("9",OSIP_codeframe_private_unique$school_zip),]

#how many OSIP codeframe public school codes not in WA DOH codeframe?
OSIP_codeframe_public_unique<- subset(OSIP_codeframe_public, OSIP_codeframe_public$school_code %in% WA_DOH_codeframe$school_code == F)


#add placeholder disrict code to private unique
OSIP_codeframe_private_unique$district_code <- 8675309

#assemble final codeframe
final_codeframe <- rbind(WA_DOH_codeframe,OSIP_codeframe_private_unique,OSIP_codeframe_public_unique)

#remove duplicate values
final_codeframe <- final_codeframe[!duplicated(final_codeframe$school_code),]

```


```{r complete data sets}
#2012
#use lambda function (apply) to extract zip code 2012, note that I could have 
#used sapply or mapply here, but instead chose drop = False to work on a single
#vector (column) because I may re-use this code on multiple columns later.
#Note that I'm using a subset of some codeframes.  That is because the final codeframe does #not contain all values of school code for 2012.  This causes type(0) values, which 
#apparently forces a list instead of a vector into the DF.I use split apply combine.

#clear WA DOHH 012 zip, city, district
doh_k12_2012$school_zip <- NULL
doh_k12_2012$school_city <- NULL
doh_k12_2012$district_code <- NULL

#subset final codeframe for 2012 data
final_codeframe_subset2012 <- subset(final_codeframe, doh_k12_2012$school_code %in% final_codeframe$school_code)
doh_k12_2012_in <- subset(doh_k12_2012, doh_k12_2012$school_code %in% final_codeframe$school_code)
doh_k12_2012_out <- subset(doh_k12_2012, doh_k12_2012$school_code %in% final_codeframe$school_code == F)

#insert school zip
doh_k12_2012_in$school_zip <- apply(doh_k12_2012_in[,c("school_code"), drop = F],1, function(x) {final_codeframe$school_zip[final_codeframe$school_code == x]})


#insert school_city 2012
doh_k12_2012_in$school_city <- apply(doh_k12_2012_in[,c("school_code"), drop = F],1, function(x) {final_codeframe$school_city[final_codeframe$school_code == x]})


#insert district code 2012
doh_k12_2012_in$district_code <- apply(doh_k12_2012_in[,c("school_code"), drop = F],1, function(x) {final_codeframe$district_code[final_codeframe$school_code == x]})


#insert district code 2013
doh_k12_2013$district_code <- apply(doh_k12_2013[,c("school_code"), drop = F],1, function(x) {final_codeframe$district_code[final_codeframe$school_code == x]})

#insert placeholders into WA Doh 2012 out
doh_k12_2012_out$school_zip <- "8675309"
doh_k12_2012_out$school_city <- "placeholder"
doh_k12_2012_out$district_code <- 8675309


```
####8675309 district code placeholder####
- will handle these in final WA DOH DF
- Most instances of the 867309 placeholder can be solved by creating a key relating district name to disritct code.  I will use split, apply, combine

###Combined WA DOH dataframe###


####Assemble####

```{r assemble combined DF}

#rename 2012 column "state" --> "school_state"
names(doh_k12_2012_in)[20] <- "school_state"
names(doh_k12_2012_out)[20] <- "school_state"


#rbind
WA_DOH_combined <- rbind(doh_k12_2011,doh_k12_2013,doh_k12_2012_in, doh_k12_2012_out)


```

####Finish insertion of district codes####

```{r final district code insert}
#still have some hyphenated zip codes, split them
#strip zipcode at hypehn
WA_DOH_combined$school_zip <- sapply(strsplit(WA_DOH_combined$school_zip,split = "-"),"[",1)

#subset rows with and without 8675309 district code placeholder
WA_DOH_combined_8675309_dcode <- subset(WA_DOH_combined, district_code == 8675309 | is.na(district_code))
WA_DOH_combined <- subset(WA_DOH_combined, district_code != 8675309)

dim(WA_DOH_combined[is.na(WA_DOH_combined$district_code),])





#create district_name ----- disrict code DF
district_codeframe <- WA_DOH_combined[,c("school_district", "district_code")]


#remove duplicated values
district_codeframe <- district_codeframe[!duplicated(district_codeframe$school_district),]


#complete fill in of district code
WA_DOH_combined_8675309_dcode$district_code <- apply(WA_DOH_combined_8675309_dcode[,c("school_district"), drop = F],1, function(x) {district_codeframe$district_code[district_codeframe$school_district == x]})
#rbind to combined DF
WA_DOH_combined <- rbind(WA_DOH_combined,WA_DOH_combined_8675309_dcode)
rm(WA_DOH_combined_8675309_dcode)


#change fied name of exempt to total_exemptions, to avoid confusion
names(WA_DOH_combined)[14] <- "total_exemptions"
#rename year column to school_year
names(WA_DOH_combined)[1] <- "school_year"
```



###Add Fields to combined DF###

```{r add fields to combined DF}
#specific exemption percents
WA_DOH_combined$percent_exempt_DTTd <- round((WA_DOH_combined$exempt_DTTd/WA_DOH_combined$enrolled)*100,digits = 2)

WA_DOH_combined$percent_exempt_pertussis <- round((WA_DOH_combined$exempt_pertussis/WA_DOH_combined$enrolled)*100,digits = 2)

WA_DOH_combined$percent_exempt_polio <- round((WA_DOH_combined$exempt_polio/WA_DOH_combined$enrolled)*100,digits = 2)

WA_DOH_combined$percent_exempt_MMR <- round((WA_DOH_combined$exempt_MMR/WA_DOH_combined$enrolled)*100,digits = 2)

WA_DOH_combined$percent_exempt_hepB <- round((WA_DOH_combined$exempt_hepB/WA_DOH_combined$enrolled)*100,digits = 2)

WA_DOH_combined$percent_exempt_varicella <- round((WA_DOH_combined$exempt_varicella/WA_DOH_combined$enrolled)*100,digits = 2)

WA_DOH_combined$percent_exempt <- round((WA_DOH_combined$total_exemptions/WA_DOH_combined$enrolled)*100,digits = 2)


str(WA_DOH_combined)
summary(WA_DOH_combined)


```

###Convert Classes of Combined DF###

```{r convert classes of combined}
#convert categorical variables to factors
WA_DOH_combined$school_type <- as.factor(WA_DOH_combined$school_type)
WA_DOH_combined$year <- as.factor(WA_DOH_combined$school_year)
WA_DOH_combined$reported <- as.factor(WA_DOH_combined$reported)
```


###Tidy the combined DF###

```{r tidy combined DF}
#gather exemptions
WA_DOH_tidy <- gather(WA_DOH_combined,exemption_type,exemptions,medical_exempt, personal_exempt, religious_exempt)

#add percent by exemption type
WA_DOH_tidy$percent_exemption_type <- round((WA_DOH_tidy$total_exemptions/WA_DOH_tidy$enrolled)*100,digits = 2)

```

```{r for udacity coaches}
#what I started with
str(read.delim('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/2011.txt', quote="" , stringsAsFactors=FALSE))
summary(read.delim('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/2011.txt', quote="" , stringsAsFactors=FALSE))

str(read.delim('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/2012.txt', quote="" , stringsAsFactors=FALSE))
summary(read.delim('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/2012.txt', quote="" , stringsAsFactors=FALSE))

str(read.delim('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/2013.txt', quote="" , stringsAsFactors=FALSE))
summary(read.delim('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/2013.txt', quote="" , stringsAsFactors=FALSE))

#Current versions
str(WA_DOH_combined)
summary(WA_DOH_combined)

str(WA_DOH_tidy)
summary(WA_DOH_tidy)


```

