---
title: "Udacity EDA R Final Project"
author: "Andrew Thomas Bauman"
date: "Tuesday, February 10, 2015"
output: html_document
---

#Introduction#

Recent news media coverage and my own personal interest inspired me to explore a school vaccination data set for WA state.  WA DOH employees generously assembled 3 data sets for my project for school years 2011-2012, 2012-2013, and 2013-2014.  The 2011-2012 set is the most comprehensive in terms of school covered and relavant fields.  I will use the following approach:

- Align 2012 and 2013 data with 2011 data, clean, and homogenize
- Tidy the data
- Concatonate the data
- Merge with demographic data
- Explore the data

###Set Path and Load Libraries###

```{r setpath}
setwd("~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject")
```

```{r load libraries}
library(RSocrata) #to access socrata data
library(ggplot2)
library(gridExtra)
library(dplyr)
library(alr3)
library(GGally)
library(reshape2)
library(acs) #to access census data
library(tidyr) #bare bones data tidying library

```

###Load WA DOH Data Sets###

Named data sets based on starting year of school year.  That is also how the year variable was assigned.  So 2011 referese to the school year (SY) 2011-2012

```{r load school vaccination exemption data sets}

#WA DOH SY 2011-2012 K-12
doh_k12_2011<- read.csv('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/WA_Vacc_K12_2011.csv')

#WA DOH SY 2012-2013 K-12
doh_k12_2012<- read.csv('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/WA_Vacc_K12_2012.csv')

#WA doh SY 2013-2014 K-12
doh_k12_2013<- read.csv('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/WA_Vacc_K12_2013.csv')
```
###Data Audit###
Most of the audit was done through simple visual inspection.  A few things stood out already, that year was missing from 2012 and 2013, and that the 2012 data had a source column which I don't intend to use.  These were easy add and drops.  I would like to align the geographic information.  The 2012 and 2013 geo info is not as comprehensive as 2011.  I will use the school code to merge all the geo information, that way everything should align and the same strings and values should exist for all schools.  I will use this as a general strategy (referencing the 2011 info) rather than trying to homogenize in other ways.  

```{r Initial Data Audit}

```

###Align Columns to 2011 Set###
Initial cleaning etc. may be clunky and manual.  I intend to semi-automate this with a pipeline prior to project completion.

```{r align to 2011 part 1, simple}
#add Year
doh_k12_2012$Year = 2012
doh_k12_2013$Year= 2013

#drop Data.Source
doh_k12_2012$Data.Source <- NULL

#rename 2013 School Code
names(doh_k12_2013)[names(doh_k12_2013) == 'School.Building.Code'] <- 'School.Code'
```

###Align School and District information to 2011 with a Merge###
- drop school and district info columns that are indpendent of school year in 2012 and 2013 except school code
- merge subset of 2011 on district code

**Important Question**

Are there schools unique to 2012 or 2013?  I need to inspect this to determine if they are worth holding on to and getting geo info.

####School Code Audit####
Going to try subset on this.  I will subset 2012 and 2013 on those school codes != to 2011, if any.  Using the %in% operator and subsetting on False (not in 2011). 
```{r School Code Audit}
#2012 Unique School Code relative to 2011
schoolcode_2012_unique_2011<- subset(doh_k12_2012, doh_k12_2012$School.Code %in% doh_k12_2011$School.Code == F)

  
#2013 Unique School Code relative to 2011
schoolcode_2013_unique_2011<- subset(doh_k12_2013, doh_k12_2013$School.Code %in% doh_k12_2011$School.Code == F)

#2011 unique school Code relative to 2012 and 2013
schoolcode_2011_unique<- subset(doh_k12_2011, (doh_k12_2011$School.Code %in% doh_k12_2012$School.Code == F) & (doh_k12_2011$School.Code %in% doh_k12_2013$School.Code == F))


#2012 unique school code relative to 2011 and 2013
schoolcode_2012_unique<- subset(doh_k12_2012, (doh_k12_2012$School.Code %in% doh_k12_2011$School.Code == F) & (doh_k12_2012$School.Code %in% doh_k12_2013$School.Code == F))

#2013 unique school code relative to 2011 and 2012
schoolcode_2013_unique<- subset(doh_k12_2013, (doh_k12_2013$School.Code %in% doh_k12_2012$School.Code == F) & (doh_k12_2013$School.Code %in% doh_k12_2011$School.Code == F))

#How many of the unique schools are in King County?
summary(schoolcode_2012_unique$County == 'KING')
summary(schoolcode_2013_unique$County == 'KING')
summary(schoolcode_2011_unique$School.County == 'KING')

```
The 2012 data set is missing address data, but there are only 31 unique school codes in this data set, so there are only 31 missing addresses.


```{r align data to 2011 part 2}

```

