---
title: "Udacity EDA R Final Project"
author: "Andrew Thomas Bauman"
date: "Tuesday, February 10, 2015"
output: html_document
---

#Introduction#

Recent news media coverage and my own personal interest inspired me to explore a school vaccination data set for WA state.  WA DOH employees generously assembled 3 data sets for my project for school years 2011-2012, 2012-2013, and 2013-2014.  The 2011-2012 set is the most comprehensive in terms of school covered and relevant fields  Each data set contains an observation about a school.  The 3 data sets will be concantonated into one data set and then explored.  

Additonal information can be provided by US Census Bureau data and data from the Office of the Superintendent of Public Instruction (OSPI) in Washington State.  This includes geographic data, demograhic data, and economic data at multiple levels of aggregation.  Final data frames will likely include:

- school vaccination data frame for 3 school years (2011-2012,2012-2013,2013-2014)
- Demographic data frame at the school, district, or county level
- school profile data frame
- economic data frame at the zip code level

I will use the following approach to assemble my initial data frame and explore the data:

* Read and audit each of the 3 school vaccination data frames
* Choose columns for the final school vaccination data frame  
* Prepare the 3 data frames to match item 2
* Concatonate the data
* Tidy the data
* Explore the data
* Merge with relevant demographic, economic, or profile data
* Explore the merged data

##Prepare the R Environment##

###Set Path and Load Libraries###

```{r setpath}
setwd("~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject")
```

```{r load libraries}
library(RSocrata) #to access socrata data
library(ggplot2)
library(gridExtra)
library(dplyr)
library(alr3)
library(GGally)
library(reshape2)
library(acs) #to access census data
library(tidyr) #bare bones data tidying library

```
##Read and Audit WA DOH K-12 School Vaccination Data Frames##

###Read WA DOH Data Sets###

Named data sets based on starting year of school year.  That is also how the year variable was assigned.  So 2011 referes to the school year (SY) 2011-2012

```{r load school vaccination exemption data sets}

#WA DOH SY 2011-2012 K-12
doh_k12_2011<- read.csv('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/WA_Vacc_K12_2011.csv')

#WA DOH SY 2012-2013 K-12
doh_k12_2012<- read.csv('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/WA_Vacc_K12_2012.csv')

#WA doh SY 2013-2014 K-12
doh_k12_2013<- read.csv('~/Udacity/dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/data/WA_Vacc_K12_2013.csv')
```
###Data Audit###
Most of the audit was done through simple visual inspection of fields and the first 20 rows of data. The columns of the dataframes don't match, there is missing information, and information was recorded in different ways.  The data is not tidy.  Tidying will be saved for the final school vaccination DF.

**Columns for the final school vaccination data frame**
I'm saving some of the geo information, combining some fields, and dropping calculated fields.

year,school_code,school_name,school_type,school_city,school_state,school_zip,school_county,school_district,district_code,grade_levels,reported,enrolled,exempt,medical_exempt,personal_exempt,religious_exempt,exempt_DTTd,exempt_pertussis,exempt_polio,exempt_MMR,exempt_hepB,exempt_varicella

```{r Initial Data Audit}

```

```{r align to 2011 part 1, simple}
#add Year
doh_k12_2012$Year = 2012
doh_k12_2013$Year= 2013

#drop Data.Source
doh_k12_2012$Data.Source <- NULL

#rename 2013 School Code
names(doh_k12_2013)[names(doh_k12_2013) == 'School.Building.Code'] <- 'School.Code'

#drop 2012 K.12.Number.Complete
doh_k12_2012$K.12.Number.Complete <- NULL

```

###Align School and District information to 2011 with a Merge###
- drop school and district info columns that are indpendent of school year in 2012 and 2013 except school code
- merge subset of 2011 on district code

**Important Question**

Are there schools unique to 2012 or 2013?  I need to inspect this to determine if they are worth holding on to and getting geo info.

####School Code Audit####
Going to try subset on this.  I will subset 2012 and 2013 on those school codes != to 2011, if any.  Using the %in% operator and subsetting on False (not in 2011). 
```{r School Code Audit}
#2012 Unique School Code relative to 2011
schoolcode_2012_unique_2011<- subset(doh_k12_2012, doh_k12_2012$School.Code %in% doh_k12_2011$School.Code == F)

  
#2013 Unique School Code relative to 2011
schoolcode_2013_unique_2011<- subset(doh_k12_2013, doh_k12_2013$School.Code %in% doh_k12_2011$School.Code == F)

#2011 unique school Code relative to 2012 and 2013
schoolcode_2011_unique<- subset(doh_k12_2011, (doh_k12_2011$School.Code %in% doh_k12_2012$School.Code == F) & (doh_k12_2011$School.Code %in% doh_k12_2013$School.Code == F))


#2012 unique school code relative to 2011 and 2013
schoolcode_2012_unique<- subset(doh_k12_2012, (doh_k12_2012$School.Code %in% doh_k12_2011$School.Code == F) & (doh_k12_2012$School.Code %in% doh_k12_2013$School.Code == F))

#2013 unique school code relative to 2011 and 2012
schoolcode_2013_unique<- subset(doh_k12_2013, (doh_k12_2013$School.Code %in% doh_k12_2012$School.Code == F) & (doh_k12_2013$School.Code %in% doh_k12_2011$School.Code == F))

#How many of the unique schools are in King County?
summary(schoolcode_2012_unique$County == 'KING')
summary(schoolcode_2013_unique$County == 'KING')
summary(schoolcode_2011_unique$School.County == 'KING')

```
The 2012 data set is missing address data, but there are only 31 unique school codes in this data set, so there are only 31 missing addresses.

Going to rename the 2011 columns.



```{r rename columns}
colnames_2011 = c('year','school_code','school_name','school_type','school_street','school_city','school_state','school_zip','school_county','school_district','district_code','grade_levels','reported','enrolled','exempt','percent_exempt','medical_exempt','percent_medical_exempt','personal_exempt','percent_personal_exempt','religious_exempt','percent_religious_exempt','exempt_DTTd','exempt_pertussis','exempt_polio','exempt_measles','exempt_mumps','exempt_rubella','exempt_hepB','exempt_varicella')

colnames(doh_k12_2011) <- colnames_2011

colnames_2012 = c('school_code','school_name','school_type','school_district','school_county','reported','grade_levels','enrolled','exempt','medical_exempt','personal_exempt','religious_exempt','religious_member_exempt','exempt_DTTd','exempt_pertussis','exempt_polio','exempt_MMR','exempt_hepB','exempt_varicella','year')

colnames(doh_k12_2012) <- colnames_2012


```
The 2011 data set contains a field for each component of MMR.  I suspect these are all the same value since I'm not aware of administering MMR components seperately as a common practice.  Are these the same?  If so, I'm going to combine them into just MMR.

```{r more audit}
summary(doh_k12_2011$exempt_measles == doh_k12_2011$exempt_mumps)
summary(doh_k12_2011$exempt_measles == doh_k12_2011$exempt_rubella)
summary(doh_k12_2011$exempt_mumps == doh_k12_2011$exempt_rubella)

```
Hmm, it looks like the values differ in some instances, let's inspect that a bit.

```{r inspect 2011 MMR}
summary(doh_k12_2011$exempt_measles)
summary(doh_k12_2011$exempt_mumps)
summary(doh_k12_2011$exempt_rubella)

reported_measles = doh_k12_2011$exempt_measles[doh_k12_2011$reported == 'Yes']
reported_mumps = doh_k12_2011$exempt_mumps[doh_k12_2011$reported == 'Yes']
reported_rubella = doh_k12_2011$exempt_rubella[doh_k12_2011$reported == 'Yes']

summary(reported_measles - reported_rubella)
summary(reported_measles - reported_mumps)
summary(reported_mumps - reported_rubella)
doh_k12_2011[max(reported_measles-reported_rubella),]
```
I don't understand some of the values the summaries are giving.  It looks like in most or all instances that M, M, and R are the same.  I'm going to use measles as MMR.  I will renamed the column and drop the other two.

```{r 2011 drop and rename}
#drop mumps and rubella
doh_k12_2011$exempt_mumps <- NULL
doh_k12_2011$exempt_rubella <- NULL
colnames(doh_k12_2011)
colnames(doh_k12_2012)
#rename measles to MMR
names(doh_k12_2011)[names(doh_k12_2011) == 'exempt_measles'] <- 'exempt_MMR'
```
For 2012 it looks like religious exempt and religous member exempt are recording the same type of information, but seperately (exemptions due to religion).  I'm going to combine values onto $religious_exempt.

```{r 2012 combine religious}
#combine religious exemptions
doh_k12_2011$religious_exempt[]
```

```{r align data to 2011 part 2}

```

