---
title: "Exploration of WA DOH School Vaccine Exemption Data for 2011-2013 School Year"
author: "Andrew Thomas Bauman"
date: "Saturday, March 14, 2015"
output: html_document
---
#Introduction#

Recent news media coverage and my own personal interest inspired me to explore a school vaccination data set for WA state.  WA DOH employees generously assembled 3 data sets for my project for school years 2011-2012, 2012-2013, and 2013-2014.  The 2011-2012 set is the most comprehensive in terms of school covered and relevant fields  Each data set contains an observation about a school.  The 3 data sets were bound into one data set, then explored.  

The initial data and cleaning/assembly of the combined data frame can be found here:
[githubrepo](https://github.com/baumanab/EDA_R_Udacity/tree/master/EDA_R_FinalProject)

The current state is a rough approach and far from perfect.  I plan to revisit this as I discover additional elements that need cleaning.  I am currently working on a more readable and organized version of the data wrangling which involves piping/chaining the code together, so that changes are easy to make and push to the .rds file called as data for this project.

**Disclaimer:**
While I am a trained scientist, I am not an expert statistician or programmer.  This is a student project.  My data cleaning and wrangling is not perfect.  As such, please exercise caution and good judgement regarding my observations and conclusions.  **Feedback on any aspect of this project is encouraged and most welcome.**

#Prepare Environment and Load Data#

```{r working directory, .RData I/O}
setwd("Udacity//dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/")
save.image("project_pause.RData")
load("project_pause.RData")
```

```{r data load}
#load current rds
vacc <- readRDS('data/WA_DOH_Combined.rds')
```


```{r library load}
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(GGally)
library(reshape2)
library(tidyr)

```

#Exploratory Data Analysis#

##Initial Survey##

```{r summary, structure, names}
str(vacc)
summary(vacc)
names(vacc)

```
###Who is and is not reporting###
Many schools have not been reporting the vaccination status of their students.  Non-reporting is responsible for many of the Na values in this data set.  Who is and is not reporting?  Is this changing from year to year?

```{r reported yes VS no}
table(vacc$reported)
with(vacc,table(school_year, reported))
with(vacc,table(school_year, school_type, reported == "No"))
prop.table(with(vacc,table(school_year== "2011", school_type, reported)))*100

```
The 2011 data set has the most not reported.  The 2012 data set has 0 not reported.  This seems unusual given the number of not reported in the bracketing school years.  I wonder if the data provided in 2012 was only from schools that reported.  I do note that the 2012 data set has the most amount of schools as well.  Interesting, I will follow up on this.  2013 has 302 schools not reporting.  The data set actually didn't contain this information, and I surmised it for Na values in other columns (see data wrangling RMD).  Several news sources have reported that ~300 schools are failing to report.  That doesn't assure the accuracy of my calculation, but it is consistent.



