---
title: "Exploration of WA DOH School Vaccine Exemption Data for 2011-2013 School Year"
author: "Andrew Thomas Bauman"
date: "Saturday, March 14, 2015"
output: html_document
---
#Introduction#

Recent news media coverage and my own personal interest inspired me to explore a school vaccination data set for WA state.  WA DOH employees generously assembled 3 data sets for my project for school years 2011-2012, 2012-2013, and 2013-2014.  The 2011-2012 set is the most comprehensive in terms of school covered and relevant fields  Each data set contains an observation about a school.  The 3 data sets were bound into one data set, then explored.  

The initial data and cleaning/assembly of the combined data frame can be found here:
[githubrepo](https://github.com/baumanab/EDA_R_Udacity/tree/master/EDA_R_FinalProject)

The current state is a rough approach and far from perfect.  I plan to revisit this as I discover additional elements that need cleaning.  I am currently working on a more readable and organized version of the data wrangling which involves piping/chaining the code together, so that changes are easy to make and push to the .rds file called as data for this project.

**Disclaimer:**
While I am a trained scientist, I am not an expert statistician or programmer.  This is a student project.  My data cleaning and wrangling is not perfect.  As such, please exercise caution and good judgement regarding my observations and conclusions.  **Feedback on any aspect of this project is encouraged and most welcome.**

#Prepare Environment and Load Data#

```{r working directory, .RData I/O}
setwd("Udacity//dataAnalysisR/EDA_R_Udacity/EDA_R_FinalProject/")
#save.image("project_pause.RData")
load("project_pause.RData")

```

```{r data load}
#load current rds
vacc <- readRDS('data/WA_DOH_Combined.rds')
```


```{r library load}
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(GGally)
library(reshape2)
library(tidyr)

```

#Exploratory Data Analysis#

##Initial Survey##

```{r summary, structure, names}
str(vacc)
summary(vacc)
names(vacc)

```
###Who is and is not reporting###
Many schools have not been reporting the vaccination status of their students.  Non-reporting is responsible for many of the Na values in this data set.  Who is and is not reporting?  Is this changing from year to year?

```{r reported yes VS no}
#counts of reported
table(vacc$reported)
#counts of repored for each year
with(vacc,table(school_year, reported))[,2:3]
#counts of reported for each school type
with(vacc,table(school_type, reported))[,2:3]
#counts of reported for each school type for each school year
with(vacc,table(school_type, reported, school_year))
#percent reported for each school type
prop.table(with(vacc,table(school_type, reported)),1)[,2:3]*100
#percent reported for each school type for each year
prop.table(with(vacc,table(school_type, reported,school_year == "2011")),c(1,3))[,2:3,2]*100
prop.table(with(vacc,table(school_type, reported,school_year == "2012")),c(1,3))[,2:3,2]*100
prop.table(with(vacc,table(school_type, reported,school_year == "2013")),c(1,3))[,2:3,2]*100

```
The 2011 data set has the most not reported.  The 2012 data set has 0 not reported.  This seems unusual given the number of not reported in the bracketing school years.  I wonder if the data provided in 2012 was only from schools that reported.  I do note that the 2012 data set has the most amount of schools as well.  Interesting, I will follow up on this.  2013 has 302 schools not reporting.  The data set actually didn't contain this information, and I surmised it for Na values in other columns (see data wrangling RMD).  Several news sources have reported that ~300 schools are failing to report.  That doesn't assure the accuracy of my calculation, but it is consistent.

```{r vacc object derivatives}
#subset on reported == "Yes"
vaccReported <- subset(vacc, vacc$reported == "Yes")
names(vaccReported)
#create column for nonmedical_exempt
vaccReported$nonmedical_exempt <- vaccReported$religious_exempt + vaccReported$personal_exempt
#add percent_nonmedexempt
vaccReported$percent_nonmedexempt <- (vaccReported$nonmedical_exempt/vaccReported$enrolled) *100
#change all schools that are not public to not_Public
vaccReported$school_type <- as.character(vaccReported$school_type)
vaccReported$school_type[vaccReported$school_type != "Public"] <- "not_Public"
unique(vaccReported$school_type)
vaccReported$school_type <- as.factor(vaccReported$school_type)

#remove schools where exemption counts don't reconcile with eachother or enrollment
#I noticed this was the case during some of my intitial plotting of the data
#everything checks out except for nonmedical exempt vs total enrolled and nonmedical exempt + medical exempt vs. total exempt.  It is the same 2 observations in each case

vaccReported <- vaccReported[vaccReported$nonmedical_exempt  <= vaccReported$enrolled,]

#while exploring population distributions, noticed that 10 rows were reported = Yes, but with an enrollment of 0.  Removing these rows.

vaccReported <- vaccReported[vaccReported$enrolled != 0,]


#noticed in my exploration that there was one very big school (10,700 students).  This seems unrealistic for a town the size of Kelso.  That would be ~2600 studens per class.  Some searching puts the number at 1819.  
vaccReported[vaccReported$enrolled > 10000,][3]
vaccReported[vaccReported$school_name == "Kelso High School",][13]

#school digger data puts the 2013 enrollment at 1562

http://www.schooldigger.com/go/WA/schools/0000300593/school.aspx

#correct kelso high school enrollment
vaccReported[vaccReported$school_name == "Kelso High School",][13][3,] <- 1562

#save vaccReported as RDS
saveRDS(vaccReported, "data/vaccReported")


#create subset of schools with populations of at least 100
vaccReported_NLT100 <- subset(vaccReported,vaccReported$enrolled >= 100)

#see enrollment distributions below, the high enrollment data is suspect.  Creating a subset of enrollment from 100 to 2,500

vaccReported_100To2500 <- subset(vaccReported_NLT100, vaccReported_NLT100$enrolled <=2500)








```
```{r distributions}

#summary of enrollment
summary(vaccReported$enrolled)
by(vaccReported$enrolled, vaccReported$school_year,summary)

#these numbers seem unrealistic to me, based on the grade levels the enrollment covers.  I'm going to proceed with the data analysis, but take the level of enrollment with a grain of salt.
vaccReported[vaccReported$enrolled == 5076,][11]
vaccReported[vaccReported$enrolled == 3383,][11]
vaccReported[vaccReported$enrolled == 3416,][11]

ggplot(aes(x = enrolled), data = vaccReported) + 
  geom_histogram(binwidth = 30,  color = 'black',fill = '#F79420') +
  scale_x_continuous(limits = c(0,5100), breaks = seq(0, 5100, 500))

ggplot(aes(x = enrolled), data = vaccReported) + 
  geom_histogram(binwidth = .1,  color = 'black',fill = '#F79420') +
  scale_x_log10()



#decided to create subset of data with students from 100 to 2500
summary(vaccReported_100To2500$enrolled)
by(vaccReported_100To2500$enrolled, vaccReported_100To2500$school_year,summary)

ggplot(aes(x = enrolled), data = vaccReported_100To2500) + 
  geom_histogram(binwidth = 30,  color = 'black',fill = '#F79420') +
  scale_x_continuous(limits = c(0,2500), breaks = seq(0, 2500, 500))

ggplot(aes(x = enrolled), data = vaccReported_100To2500) + 
  geom_histogram(binwidth = .1,  color = 'black',fill = '#F79420') +
  scale_x_log10()


#summary of total exemptions
summary(vaccReported$total_exemptions)
by(vaccReported$total_exemptions, vaccReported$school_year,summary)

#histograms of total exemptions
ggplot(aes(x = total_exemptions), data = vaccReported) + 
  geom_histogram(binwidth = 5,  color = 'black',fill = '#F79420') +
  scale_x_continuous(limits = c(0,250), breaks = seq(0, 250, 25))

#total exempt by year
ggplot(aes(x = total_exemptions), data = vaccReported) + 
  geom_histogram(binwidth = 5,  color = 'black',fill = '#F79420') +
  facet_wrap(~school_year) +
  scale_x_continuous(limits = c(0,250), breaks = seq(0, 250, 25))

#log scale (longtail data)
ggplot(aes(x = total_exemptions), data = vaccReported) + 
  geom_histogram(binwidth = .1,  color = 'black',fill = '#F79420') +
  facet_wrap(~school_year) +
  scale_x_log10() +
  xlab("Log10(Total Exemptions)")

ggplot(aes(x = total_exemptions), data = subset(vaccReported, total_exemptions >0)) + 
  geom_histogram(binwidth = .1,  color = 'black',fill = 'red') +
  scale_x_log10() +
  xlab("Log10(Total Exemptions)")

#explore medical exemptions
summary(vaccReported$medical_exempt)
by(vaccReported$medical_exempt, vaccReported$school_year,summary)
#medical exempt
ggplot(aes(x = medical_exempt), data = vaccReported) + 
  geom_histogram(binwidth = .1,  color = 'black',fill = '#F79420') +
  facet_wrap(~school_year) +
  scale_x_log10() +
  xlab("Log10(medical_exemptions)")

#explore non-medical exemptions
summary(vaccReported$nonmedical_exempt)
by(vaccReported$nonmedical_exempt, vaccReported$school_year,summary)
#non-medical exempt
ggplot(aes(x = nonmedical_exempt), data = vaccReported) + 
  geom_histogram(binwidth = .1,  color = 'black',fill = '#F79420') +
  facet_wrap(~school_year) +
  scale_x_log10() +
  xlab("Log10(nonMedical_exemptions)")





ggplot(aes(x = percent_exempt, y = ..count../sum(..count..)), data = vaccReported) +
  geom_freqpoly(binwidth = .1, aes(color = school_year)) +
  facet_wrap(~school_county) +
  scale_x_log10()

ggplot(aes(x = percent_exempt, y = ..count../sum(..count..)), data = vaccReported) +
  geom_freqpoly(binwidth = .1, aes(color = school_year)) +
  facet_wrap(~school_county, scales = "free_y") +
  scale_x_log10()



```


```{r sandbox}

##freqpoly proportion of percent exempt, by county, color = year
ggplot(aes(x = percent_exempt, y = ..count../sum(..count..)), data = vaccReported) +
  geom_freqpoly(binwidth = .5, aes(color = school_year)) +
  facet_wrap(~school_county) +
  scale_x_log10()

##freqpoly density of percent exempt, by county, color = year
ggplot(aes(x = percent_exempt, y = ..density..), data = vaccReported) +
  geom_freqpoly(binwidth = .5, aes(color = school_year)) +
  facet_wrap(~school_county) +
  scale_x_log10()

##top 200
top200_allyears_allpop <- arrange(vaccReported, desc(percent_exempt))[1:200,c("school_year","school_name", "school_county", "school_city","enrolled", "total_exemptions", "percent_exempt")]

top200_allyears_NLT100 <- arrange(subset(vaccReported, vaccReported$enrolled >= 100), desc(percent_exempt))[1:200,c("school_year","school_name", "school_county", "school_city","enrolled", "total_exemptions", "percent_exempt")]



state_rate <- mean(vaccReported$percent_exempt, na.rm=T)
state_rate_NLT100 <- mean(vaccReported_NLT100$percent_exempt, na.rm = T)

summarize(vaccReported_NLT100, Mean_percent_exempt = mean(percent_exempt, na.rm = TRUE))

summarize(vaccReported_NLT100, Mean_percent_nonmedexempt = mean(percent_nonmedexempt, na.rm = TRUE))

#boxplot to compare means/medians NLT100 vs reg



##total percent exempt all schools

##state level by year
ggplot(vaccReported, aes(x = school_state, y = percent_exempt)) +
  geom_boxplot() +
  geom_jitter(alpha = .5, aes(size = enrolled, color = enrolled)) +
  facet_wrap(~school_year) +
  scale_color_gradientn(colours = c("red", "orange", "yellow")) #fix color scheme and do plots where outier pops are removed

#year level by county
ggplot(vaccReported, aes(x = school_year, y = percent_exempt)) +
  geom_boxplot(outlier.colour = "red") +
  facet_wrap(~school_county)

##total percent exempt all schools

##state level by year
ggplot(vaccReported_NLT100, aes(x = school_state, y = percent_exempt)) +
  geom_boxplot() +
  geom_jitter(alpha = .5, aes(size = enrolled, color = enrolled)) +
  facet_wrap(~school_year) +
  scale_color_gradientn(colours = c("red", "orange", "yellow")) #fix color scheme and do plots where outier pops are removed

##year level by county
ggplot(vaccReported_NLT100, aes(x = school_year, y = percent_exempt)) +
  geom_boxplot(outlier.colour = "red") +
  facet_wrap(~school_county)

#nonmedical exempt

#
ggplot(vaccReported_NLT100, aes(x = school_state, y = percent_nonmedexempt)) +
  geom_boxplot(outlier.colour = "red") +
  facet_wrap(~school_year)

#
ggplot(vaccReported_NLT100, aes(x = school_year, y = percent_nonmedexempt)) +
  geom_boxplot(outlier.colour = "red") +
  facet_wrap(~school_county)


#does cutting school pops have much effect on the mean and median, if not, go with the NLT100

#does setting an uppler limit on students have much effect on median (getting ride of huge school)

#might actually want to take a look at the distribution of enrollment and form an approach


#what proportion of total exemptions are personal exemptions




#should analyze to see if there is correlation between studdents enrolled and percent exempt

ggplot(vaccReported,aes(x = enrolled, y = percent_exempt)) +
  geom_point(alpha = .1, aes(color = school_year)) +
  geom_smooth()

ggplot(subset(vaccReported, vaccReported$enrolled <= 1000),aes(x = enrolled, y = percent_exempt)) +
  geom_point(alpha = .25, aes(color = school_year)) +
  geom_smooth()





### By Pub/Private
by.type <- data.sub %>% group_by(Type) %>% summarize(Mean.PBE = round(mean(Exempt, na.rm=TRUE), 2), Students=sum(enrollment, na.rm = TRUE))


### by specific type
by.mwc <- data.sub %>%  group_by(MWC) %>% summarize(Mean.PBE = round(mean(Exempt, na.rm=TRUE), 2), Median.PBE = round(median(Exempt, na.rm=TRUE), 2), Max.PBE = round(max(Exempt, na.rm=TRUE), 2), Min.PBE = round(min(Exempt, na.rm=TRUE), 2), Schools=n(), Students=sum(enrollment, na.rm=TRUE)) %>% na.omit() %>% arrange(desc(Mean.PBE))

### By County
by.county <- data.sub %>% group_by(county) %>% summarize(Mean.PBE = round(mean(Exempt, na.rm=TRUE), 2), Students=sum(enrollment, na.rm=TRUE)) %>% na.omit() %>% arrange(desc(Mean.PBE))

### By School within County
school.by.county <- data.sub %>% group_by(county, name) %>% summarize(Mean.PBE = round(mean(Exempt, na.rm=TRUE), 2), Students=sum(enrollment, na.rm=TRUE)) %>% na.omit() %>% arrange(desc(Mean.PBE))

### By City
by.city <- data.sub %>% group_by(city) %>% summarize(Mean.PBE = round(mean(Exempt, na.rm=TRUE), 2), Students=sum(enrollment, na.rm=TRUE)) %>% na.omit() %>% arrange(desc(Mean.PBE))

### By School within City
school.by.city <- data.sub %>% group_by(city, code) %>% summarize(Mean.PBE = round(mean(Exempt, na.rm=TRUE), 2), Students=sum(enrollment, na.rm=TRUE)) %>% na.omit() %>% arrange(desc(Mean.PBE))

### By District
by.district <- data.sub %>% group_by(district) %>% summarize(Mean.PBE = round(mean(Exempt, na.rm=TRUE), 2), Students=sum(enrollment, na.rm=TRUE)) %>% na.omit() %>% arrange(desc(Mean.PBE))

### By School within District
school.by.district <-  data.sub %>% group_by(district, code) %>% summarize(Mean.PBE = round(mean(Exempt, na.rm=TRUE), 2), Students=sum(enrollment, na.rm=TRUE)) %>% na.omit() %>% arrange(desc(Mean.PBE))

### By School within Type
school.by.type <- data.sub %>% group_by(MWC, name) %>% summarize(Mean.PBE = round(mean(Exempt, na.rm=TRUE), 2), Students=sum(enrollment, na.rm=TRUE)) %>% na.omit() %>% arrange(desc(Mean.PBE))

filter(school.by.type, MWC=="Private Catholic") # Top Catholic PBE is
                                        # a homeschool service


### County Level
p <- ggplot(by.county, aes(x=log(Students), y=Mean.PBE))
p1 <- p + geom_point(alpha=0.6) + theme_bw() +
    ylab("Percent of Students with a Personal Belief Exemption") +
        xlab("log N Students in County \n") +  ggtitle("Kindergarten Vaccine Exemption Rates in California, County Level") + scale_color_manual(values=cb.palette)

### Pick out some outliers
ind <- with(by.county, (Mean.PBE>2*IQR(Mean.PBE)))
data.out <- droplevels(by.county[ind,])

pdf(file="figures/pbe-by-county.pdf", height=5, width=8)
p2 <- p1 + geom_text(data = data.out, aes(x=log(Students),
                         y=Mean.PBE, label=county), hjust=-0.1, size=2)
print(p2)
credit("Data: CA Dept of Public Health, 2015. Kieran Healy: http://kieranhealy.org")
dev.off()

ggsave(
    "figures/pbe-by-county.png",
    p2,
    width=8,
    height=5,
    dpi=300
    )

```



