Lesson 5
========================================================

###Load dependencies, data, and set path###

```{r libraries and path}
setwd("~/Udacity/dataAnalysisR/EDA_R_Udacity/lesson5")
getwd()
list.files()

library(ggplot2)
library(gridExtra)
library(dplyr)
library('alr3')

pf <- read.csv("~/Udacity/dataAnalysisR/EDA_R_Udacity/lesson3/pseudo_facebook.tsv", sep = '\t')
names(pf)

```

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + geom_boxplot()

ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + geom_histogram(binwidth = .5,stat = 'identity')

# Write code to create a new data frame,
# called 'pf.fc_by_age_gender', that contains
# information on each age AND gender group.

# The data frame should contain the following variables:

#    mean_friend_count,
#    median_friend_count,
#    n (the number of users in each age and gender grouping)

# Here is an example of the structure of your data frame. Your
# data values will be different.

#   age gender mean_friend_count median_friend_count    n
# 1  13 female          247.2953                 150  207
# 2  13   male          184.2342                  61  265
# 3  14 female          329.1938                 245  834
# 4  14   male          157.1204                  88 1201

# See the Instructor Note for two hints.

# DO NOT DELETE THESE NEXT TWO LINES OF CODE
# ==============================================================
#pf <- read.delim('/datasets/ud651/pseudo_facebook.tsv')
#suppressMessages(library(dplyr))

# ENTER YOUR CODE BELOW THIS LINE.
# ==============================================================

pf.fc_by_age_gender <- pf %>%
  group_by(age, gender) %>%
  filter(!is.na(gender))%>%
  summarise(mean_friend_count = mean(friend_count),
            median_friend_count = median(as.numeric(friend_count)),
            n = n()) %>%
  ungroup() %>% #summarise removes one layer of grouping the last group it collapsed, in this case gender, use ungroup(age) to remove grouping by age
  arrange(age)

head(pf.fc_by_age_gender)



```

***

### Plotting Conditional Summaries
Notes:  

```{r Plotting Conditional Summaries}
# Create a line graph showing the
# median friend count over the ages
# for each gender. Be sure to use
# the data frame you just created,
# pf.fc_by_age_gender.

# See the Instructor Notes for a hint.

# This assignment is not graded and
# will be marked as correct when you submit.

# ENTER YOUR CODE BELOW THIS LINE
# =================================================
ggplot(aes(x = age, y = median_friend_count), data = pf.fc_by_age_gender) +
  geom_line(aes(color = gender))
```

***

### Thinking in Ratios
Notes:  Looking at the data we see that for younger users females tend to have more friends than males.  How many?  is it twice as many, 10x??  To answer this question we need to reshape the data.

***

### Wide and Long Format
Notes:  The data is currently in long format.  We will need the data in wide format to conduct our analysis.  we are going to use the reshape2 package.

**Instructor notes:**

You can also restructure the data using the tidyr package. You can review examples and how to use the package in the Data Wrangling with R pdf.

The code to change the data frame from long format to wide (or tidy format) is shown below. I encourage you to read the Data Wrangling pdf and write code using the tidyr package before looking at the solution below. 

***

### Reshaping Data
Notes:

```{r reshape2}
#install.packages('reshape2')
library(reshape2)

pf.fc_by_age_gender_wide <- dcast(pf.fc_by_age_gender,
                                  age ~ gender,
                                  value.var = 'median_friend_count')

head(pf.fc_by_age_gender_wide)
```
Instructor Notes

It's important to use quotes around the variable name that is assigned tovalue.var. 

We could also create a similar data frame using the dplyr package. 
pf.fc_by_age_gender.wide <- pf.fc_by_age_gender %.% 
  group_by(age) %.% 
  summarise(male = friend_count.median[gender = 'male'], 
                      female = friend_count.median[gender = 'female'], 
                      ratio = female / male) %.% 
  arrange(age) 

head(pf.fc_by_age_gender.wide)

***

### Ratio Plot
Notes:

```{r Ratio Plot}
# Plot the ratio of the female to male median
# friend counts using the data frame
# pf.fc_by_age_gender.wide.

# Think about what geom you should use.
# Add a horizontal line to the plot with
# a y intercept of 1, which will be the
# base line. Look up the documentation
# for geom_hline to do that. Use the parameter
# linetype in geom_hline to make the
# line dashed.

# The linetype parameter can take the values 0-6:
# 0 = blank, 1 = solid, 2 = dashed
# 3 = dotted, 4 = dotdash, 5 = longdash
# 6 = twodash

# This assignment is not graded and
# will be marked as correct when you submit.

# ENTER YOUR CODE BELOW THIS LINE
# =================================================
ggplot(aes(x = age, y = female/male), data = pf.fc_by_age_gender_wide) +
  geom_line() +
  geom_hline(yintercept = 1, alpha = 0.3, linetype = 2)
```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}
# Create a variable called year_joined
# in the pf data frame using the variable
# tenure and 2014 as the reference year.

# The variable year joined should contain the year
# that a user joined facebook.

# See the Instructor Notes for three hints if you get
# stuck. Scroll down slowly to see one hint at a time
# if you would like some guidance.

# This programming exercise WILL BE automatically graded.

# DO NOT ALTER THE CODE BELOW THIS LINE
# ========================================================
#pf <- read.delim('/datasets/ud651/pseudo_facebook.tsv')

# ENTER YOUR CODE BELOW THIS LINE.
# ========================================================

```
pf$year_joined <- floor(2014 - pf$tenure/365)


***

### Cut a Variable
Notes:  Cut function is useful to make discrete variables from continuous or numerical variables.

Cut function information : http://www.r-bloggers.com/r-function-of-the-day-cut-2/

```{r Cut a Variable}
summary(pf$year_joined)

table(pf$year_joined)

# Create a new variable in the data frame
# called year_joined.bucket by using
# the cut function on the variable year_joined.

# You need to create the following buckets for the
# new variable, year_joined.bucket

#        (2004, 2009]
#        (2009, 2011]
#        (2011, 2012]
#        (2012, 2014]

# Note that a parenthesis means exclude the year and a
# bracket means include the year.

# Look up the documentation for cut or try the link
# in the Instructor Notes to accomplish this task.

# DO NOT DELETE THE TWO LINES OF CODE BELOW THIS LINE
# ========================================================================
#pf <- read.delim('/datasets/ud651/pseudo_facebook.tsv')
#pf$year_joined <- floor(2014 - pf$tenure / 365)

# ENTER YOUR CODE BELOW THIS LINE
# ========================================================================
pf$year_joined.bucket <- cut(pf$year_joined, c(2004,2009,2011,2012,2014))
table(pf$year_joined.bucket)
```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}
# Create a line graph of friend_count vs. age
# so that each year_joined.bucket is a line
# tracking the median user friend_count across
# age. This means you should have four different
# lines on your plot.

# You should subset the data to exclude the users
# whose year_joined.bucket is NA.

# If you need a hint, see the Instructor Notes.

# This assignment is not graded and
# will be marked as correct when you submit.

# ENTER YOUR CODE BELOW THIS LINE
# ===================================================
ggplot(aes(x = age, y = friend_count), 
              data = subset(pf, !is.na(pf$year_joined))) + 
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = median)
```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}
# Write code to do the following:

# (1) Add another geom_line to code below
# to plot the grand mean of the friend count vs age.

# (2) Exclude any users whose year_joined.bucket is NA.

# (3) Use a different line type for the grand mean.

# As a reminder, the parameter linetype can take the values 0-6:

# 0 = blank, 1 = solid, 2 = dashed
# 3 = dotted, 4 = dotdash, 5 = longdash
# 6 = twodash

# This assignment is not graded and
# will be marked as correct when you submit.

# The code from the last programming exercise should
# be your starter code!

# ENTER YOUR CODE BELOW THIS LINE
# ==================================================================
ggplot(aes(x = age, y = friend_count), 
              data = subset(pf, !is.na(pf$year_joined))) + 
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary',fun.y = mean, linetype = 3)
```

***

### Friending Rate
Notes:

```{r Friending Rate}

```

***

### Friendships Initiated
Notes:

What is the median friend rate?

What is the maximum friend rate?

```{r Friendships Initiated}

```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}

```

***

### Number of Purchases
Notes:

```{r Number of Purchases}

```

***

### Prices over Time
Notes:

```{r Prices over Time}

```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}

```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:

***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

